---
title: "Bolzenkötter"
format: html
---

```{r}
library(mlts)
library(tidyverse)

# new dataset:
# Link to OSF: https://osf.io/z2e83/
# Link to Article: https://doi.org/10.1007/s12671-024-02350-5 
# Link to Codebook: https://osf.io/z2e83/files/uph2q

# Link to ESM data on OSF
esm_link = "https://osf.io/q4czt"
survey_link = "https://osf.io/ecsj9"

# obtain datasets directly from OSF
library(osfr)

# ESM DATA 
downloaded <- osf_download(osf_retrieve_file(esm_link), path = tempdir(), conflicts = "overwrite")
esm   <- read.csv(downloaded$local_path)

# SURVEY DATA 
downloaded <- osf_download(osf_retrieve_file(survey_link), path = tempdir(), conflicts = "overwrite")
survey   <- read.csv(downloaded$local_path)

# clean-up
unlink(downloaded)
rm(downloaded)

# select relevant survey items to add to esm data 
use = c("id", "group", "pre_rtq_sum", "post_rtq_sum")
survey = survey[,use]
survey$pre_rtq_sum_cen = survey$pre_rtq_sum - mean(survey$pre_rtq_sum)
survey$post_rtq_sum_cen = survey$post_rtq_sum - mean(survey$post_rtq_sum)


# Merge data 
bolzen <- merge(x = survey, y = esm, by = c("id", "group"), all = TRUE)

```


```{r}
# inspect dataset characteristics 
length(table(bolzen$id))  # N = 100
hist(table(bolzen$id))    
table(survey$group) # 50 GI = actice control group / 50 DM = detached mindfulness
table(esm$id, esm$phase)

# esm items
esm_items = c("gluecklich","froehlich", "entspannt",  "energiegeladen",
  "traurig",  "niedergeschlagen", "aengstlich", "nervoes",
  "veraergert", "reizbar", "posaff", "negaff", "rnt")

# 
summary(bolzen[,esm_items])
# > No missings on ESM items

# additional preprocessing:
bolzen$phase_bi = ifelse(bolzen$phase == "b", 0,1)
bolzen$group_bi = ifelse(bolzen$group == "gi", 0, 1)

bolzen %>%
  pivot_longer(cols = esm_items) %>%
  ggplot(aes(x = value)) +
  geom_histogram() +
  facet_wrap(~name)
```

### Sampling Scheme 
```{r, message=F, warning=F}
#| fig-width: 12
#| fig-height: 3

bolzen$time_window = factor(bolzen$time_window, 
                            levels = c("mor", "mid", "eve"))
bolzen$timepoint = factor(bolzen$timepoint, 
                            levels = c("t0", "t1", "t2"))
bolzen$phase = factor(bolzen$phase, levels = c("b","i"))

library(ggh4x)
bolzen %>%
  filter(id %in% sample(unique(bolzen$id),size = 3)) %>%
  ggplot(aes(y = rnt, x = timepoint, group = time_window)) +
  geom_point() +
  geom_line() + 
  facet_grid2(id~phase+dayphase_counter+time_window,
              strip = strip_nested())
```


```{r}
##### WHAT WE SHOULD SEE: 
##### There are rows missing in the data. We should correct for this by adding 
##### missing data rows during estimation. 

#### start with a dummy grid that we merge to each data: 
table(bolzen$id,bolzen$day_counter)
N_days = 10
N_windows = 3
N_tp = 3

dummy_grid = data.frame(
  "day_counter"       = rep(1:N_days, each = N_windows*N_tp),
  "time_window"       = rep(c("mor", "mid", "eve"), each = N_tp, times = N_days),
  "timepoint"         = rep(c("t0","t1","t2"), times = N_days*N_windows),
  "count"             = 1:(N_days*N_windows*N_tp)
)

# add additional space for overnight-lags and between time windows
dummy_grid$shift = ifelse(dummy_grid$time_window != lag(dummy_grid$time_window),4,1)
dummy_grid$shift = ifelse(dummy_grid$day_counter > lag(dummy_grid$day_counter), 8, dummy_grid$shift)
dummy_grid$shift[1] = 0
dummy_grid$int_time_spaced = cumsum(dummy_grid$shift)

# add new grid variables to the data 
bolzen = merge(x = bolzen, y = dummy_grid, by = c("day_counter", "time_window", "timepoint"))

bolzen = bolzen %>% arrange(id, int_time_spaced)

#####
bolzen %>%
  filter(id == 1) %>%
  ggplot(aes(x = int_time_spaced, y = rnt)) +
  geom_point() +
  geom_line()

bolzen %>%
  filter(id == 1) %>%
  ggplot(aes(x = obs_counter, y = rnt)) +
  geom_point() +
  geom_line()


bolzen_spaced = create_missings(
  data = bolzen, 
  tinterval = 1,
  id = "id",
  # choose one of: 
#  time = "int_time_spaced"   # this one for higher temporal resolution
  time = "count"
  )

summary(bolzen_spaced$phase_bi)
summary(bolzen_spaced$group_bi)

# fill NAs for exogenous variables
bolzen_spaced <- bolzen_spaced %>%
  group_by(id) %>%
  mutate(
    group = unique(group[!is.na(group)]),
    pre_rtq_sum_cen = unique(pre_rtq_sum_cen[!is.na(pre_rtq_sum_cen)]),
    phase_bi = ifelse(day_counter<6, 0,1),
    group_bi = ifelse(any(group=="gi"),0,1)) %>%
  group_by(id) %>%
  fill(phase_bi, .direction = "downup") %>%
  ungroup()
  
summary(bolzen_spaced$phase_bi)
summary(bolzen_spaced$group_bi)
```

# Model Extensions 

## Multiple-Indicator Factor Model

* Steps: 
  * Check correlations between variables
  * Optional: Two-level CFA as a first step. 
  *           See how the same model looks in lavaan



### Two-level CFA
```{r}
model = mlts_model(q = 1, p = 3, max_lag = 0, fix_inno_vars = T)
mlts_paths(model)

fit = mlts_fit(model = model, 
               data = bolzen %>% filter(!(id %in% 62)),
               id = "id",
               ts = c("gluecklich", "froehlich", "entspannt"),
               na.rm = TRUE,
               iter = 2000,
               chains = 4,
               cores = 4
               )

bayesplot::mcmc_trace(fit$posteriors[,,])

summary(fit)
```

```{r}
library(lavaan)

lavmod = "
level: 1
PAw =~ gluecklich + froehlich + entspannt

level: 2
PAb =~ gluecklich + froehlich + entspannt
gluecklich ~ 0*1
PAb ~ NA*1
gluecklich ~~ 0*gluecklich
"

lavfit = cfa(model = lavmod, 
             data = bolzen %>% filter(!(id %in% 62)), 
             cluster = "id")
summary(lavfit)
```

### Three-Indicator AR(1)

* Achtung hier mit personen-spezifischen Innovationsvarianzen:

```{r}
model = mlts_model(q = 1, p = 3)

fit = mlts_fit(model = model, 
               data = bolzen %>% filter(!(id %in% 62)),
               id = "id",
               ts = c("gluecklich", "froehlich", "entspannt"),
               na.rm = TRUE,
               iter = 2000,
               chains = 4,
               cores = 4
               )

bayesplot::mcmc_trace(fit$posteriors[,,])

summary(fit)
```




























# Potential Analyses (unstructured):

## A: Compare Intervention-Phase 
* Momentary effect of intervention (t1, t2)
* linear trend 
* use type of intervention as between-level variable
* outcome: repetitive negative thinking
* With AR to begin with
```{r}
# add new variables 
bolzen$t15 = ifelse(bolzen$timepoint == "t1", 1, 0)
bolzen$t30 = ifelse(bolzen$timepoint == "t2", 1, 0)
bolzen$dayphase_counter_0 = bolzen$dayphase_counter-1

model = mlts_model(
  q = 4, 
  max_lag = 0, 
  is_exogenous = 1:3,
  incl_t0_effects = c("phi(0)_41", "phi(0)_42"),
  incl_rDSEM_effects = c("phi(s)_43"),
  inno_covs_zero = T)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("t15", "t30", "dayphase_counter_0", "rnt")
)


summary(fit)
##### VERY HIGH CORRELATION BETWEEN phi(0)_41.phi(0)_42	0.974
##### ASSUME EQUAL EFFECT for 15 and 30 minute interval: 

# ==============================================================================

# add new variables 
bolzen$intervene = ifelse(bolzen$timepoint != "t0", 1, 0)
summary(bolzen_spaced$intervene)

model = mlts_model(
  q = 3, 
  max_lag = 0, 
  is_exogenous = 1:2,
  incl_rDSEM_effects = c("phi(s)_32"),
  incl_t0_effects = c("phi(0)_31"),
  fixef_zero = c("phi(1)_32", "phi(1)_33"),
  inno_covs_zero = T)

mlts_paths(model)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("intervene", "dayphase_counter_0", "rnt")
)


summary(fit)

bayesplot::mcmc_trace(fit$posteriors)

```

* Try simple growth model:
```{r}
bolzen %>% filter(!(id %in% c(36, 51,88))) %>%
  ggplot(aes(x = dayphase_counter_0, y = rnt, group = id, color = as.factor(id))) +
  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  theme(legend.position = "none") + 
  facet_grid(group~phase)


model = mlts_model(
  q = 2, 
  max_lag = 0, 
  is_exogenous = 1,
  incl_rDSEM_effects = c("phi(s)_21"),
  ranef_pred = "group_bi",
  inno_covs_zero = T)

#mlts_paths(model)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("dayphase_counter_0", "rnt"),
  iter = 1000
)

summary(fit)
mlts_standardized(fit)
```

* Try adding the momentary effect of the mindfullness intervention
```{r}
bolzen$intervene = ifelse(bolzen$timepoint != "t0", 1, 0)

bolzen %>% filter(!(id %in% c(36, 51,88))) %>%
  ggplot(aes(x = timepoint, y = rnt)) +
#  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  geom_boxplot() +
  theme(legend.position = "none") + 
  facet_grid(group~phase)

bolzen %>% 
  group_by(id, phase, day_counter, time_window) %>%
  mutate(posaff_diff = ifelse(timepoint != "t0", posaff - posaff[timepoint=="t0"],NA)) %>%
  ggplot(aes(x = timepoint, y = posaff_diff)) +
#  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  geom_boxplot() +
  theme(legend.position = "none") + 
  facet_grid(group~phase)

bolzen %>% 
  group_by(id, phase, day_counter, time_window) %>%
  mutate(aff_diff = ifelse(timepoint != "t0", negaff - negaff[timepoint=="t0"],NA)) %>%
  ggplot(aes(x = timepoint, y = aff_diff)) +
#  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  geom_boxplot() +
  theme(legend.position = "none") + 
  facet_grid(group~phase)

bolzen %>% 
  group_by(id, phase, day_counter, time_window) %>%
  mutate(aff_diff = ifelse(timepoint != "t0", rnt - rnt[timepoint=="t0"],NA)) %>%
  ggplot(aes(x = timepoint, y = aff_diff)) +
#  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  geom_boxplot() +
  theme(legend.position = "none") + 
  facet_grid(group~phase)
```


```{r}

bolzen$intervene = ifelse(bolzen$timepoint != "t0", 1, 0)

model = mlts_model(
  q = 3, 
  max_lag = 0, 
  is_exogenous = c(1,3),
  incl_rDSEM_effects = c("phi(s)_21"),
  incl_t0_effects = c("phi(0)_23"),
  fixef_zero = c("phi(1)_23"),
  ranef_pred = "group_bi",
  inno_covs_zero = T)

#mlts_paths(model)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("dayphase_counter_0", "rnt", "intervene"),
  iter = 1000,
)

summary(fit)
mlts_standardized(fit)
```


### Testing Model: POSAFF

Are positive affect dynamics changing between baseline and 
the intervention period (irrespective of treatment)?
```{r}


model1 = mlts_model(q = 2, 
                    is_exogenous = 1, 
                    fixef_zero = "phi(1)_21",
                    incl_t0_effects = "phi(0)_21",
                    incl_interaction_effects = "phi(i)_2.1(0)2(1)",
                    fix_inno_vars = T,
                    inno_covs_zero = T)

mlts_paths(model1)

# try to estimate: 
fit1 = mlts_fit(
  model = model1, 
  data = bolzen_spaced, 
  id = "id",
  ts = c("phase_bi", "posaff"),
  iter = 2000  
  )

summary(fit1)

bayesplot::mcmc_trace(fit1$posteriors)

```

#### With group as predictor
```{r}


model = mlts_model(q = 2, 
                    is_exogenous = 1, 
                    ranef_pred = "group_bi",
                    fixef_zero = "phi(1)_21",
                    incl_t0_effects = "phi(0)_21",
                    incl_interaction_effects = "phi(i)_2.1(0)2(1)",
                    fix_inno_vars = T,
                    inno_covs_zero = T)

mlts_paths(model)

# try to estimate: 
fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, 
  id = "id",
  ts = c("phase_bi", "negaff"),
  iter = 2000,
  center_covs = F
  )

summary(fit)

bayesplot::mcmc_trace(fit1$posteriors)

```


### Testing Model: NEGAFF

Are negative affect dynamics changing between baseline and 
the intervention period (irrespective of treatment)?
```{r}


model1 = mlts_model(q = 2, 
                    is_exogenous = 1, 
                    fixef_zero = "phi(1)_21",
                    incl_t0_effects = "phi(0)_21",
                    incl_interaction_effects = "phi(i)_2.1(0)2(1)",
                    fix_inno_vars = T,
                    inno_covs_zero = T)

mlts_paths(model1)

# try to estimate: 
fit1 = mlts_fit(
  model = model1, 
  data = bolzen_spaced, 
  id = "id",
  ts = c("phase_bi", "negaff"),
  iter = 2000  
  )

summary(fit1)

bayesplot::mcmc_trace(fit1$posteriors)

```

#### With group as predictor
```{r}


model = mlts_model(q = 2, 
                    is_exogenous = 1, 
                    ranef_pred = "group_bi",
                    fixef_zero = "phi(1)_21",
                    incl_t0_effects = "phi(0)_21",
                    incl_interaction_effects = "phi(i)_2.1(0)2(1)",
                    fix_inno_vars = T,
                    inno_covs_zero = T)

mlts_paths(model)

# try to estimate: 
fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, 
  id = "id",
  ts = c("phase_bi", "negaff"),
  iter = 2000,
  center_covs = F
  )

summary(fit)
```



### Testing Model: RNT

Are negative affect dynamics changing between baseline and 
the intervention period (irrespective of treatment)?
```{r}

model1 = mlts_model(q = 2, 
                    is_exogenous = 1, 
                    fixef_zero = "phi(1)_21",
                    incl_t0_effects = "phi(0)_21",
                    incl_interaction_effects = "phi(i)_2.1(0)2(1)",
                    fix_inno_vars = T,
                    inno_covs_zero = T)

mlts_paths(model1)

# try to estimate: 
fit1 = mlts_fit(
  model = model1, 
  data = bolzen_spaced, 
  id = "id",
  ts = c("phase_bi", "rnt"),
  iter = 2000  
  )

summary(fit1)

bayesplot::mcmc_trace(fit1$posteriors)

```

#### With group as predictor
```{r}


model = mlts_model(q = 2, 
                    is_exogenous = 1, 
                    ranef_pred = "group_bi",
                    fixef_zero = "phi(1)_21",
                    incl_t0_effects = "phi(0)_21",
                    incl_interaction_effects = "phi(i)_2.1(0)2(1)",
                    fix_inno_vars = T,
                    inno_covs_zero = T)

mlts_paths(model)

# try to estimate: 
fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, 
  id = "id",
  ts = c("phase_bi", "negaff"),
  iter = 2000,
  center_covs = F
  )

summary(fit)
```





## B: Compare Intervention-Phase 
* Momentary effect of Intervation (t1, t2)
* use type of intervention as between-level variable
* outcome: repetitive negative thinking
* With AR to begin with and moderation effect successful 

```{r}
# add new variables 
bolzen$t15 = ifelse(bolzen$timepoint == "t1", 1, 0)
bolzen$t30 = ifelse(bolzen$timepoint == "t2", 1, 0)
bolzen$dayphase_counter_0 = bolzen$dayphase_counter-1

model = mlts_model(
  q = 4, 
  max_lag = 0, 
  is_exogenous = 1:3,
  incl_t0_effects = c("phi(0)_41", "phi(0)_42"),
  incl_rDSEM_effects = c("phi(s)_43"),
  inno_covs_zero = T)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("t15", "t30", "dayphase_counter_0", "rnt")
)


summary(fit)
##### VERY HIGH CORRELATION BETWEEN phi(0)_41.phi(0)_42	0.974
##### ASSUME EQUAL EFFECT for 15 and 30 minute interval: 

# ==============================================================================

# add new variables 
bolzen$intervene = ifelse(bolzen$timepoint != "t0", 1, 0)

model = mlts_model(
  q = 3, 
  max_lag = 0, 
  is_exogenous = 1:2,
  incl_rDSEM_effects = c("phi(s)_32"),
  fixef_zero = c("phi(1)_32", "phi(1)_33"),
  inno_covs_zero = T)

mlts_paths(model)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("intervene", "dayphase_counter_0", "rnt")
)


summary(fit)

bayesplot::mcmc_trace(fit$posteriors)

```

* Try simple growth model:
```{r}
bolzen %>% filter(!(id %in% c(36, 51,88))) %>%
  ggplot(aes(x = dayphase_counter_0, y = rnt, group = id, color = as.factor(id))) +
  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  theme(legend.position = "none") + 
  facet_grid(group~phase)


model = mlts_model(
  q = 2, 
  max_lag = 0, 
  is_exogenous = 1,
  incl_rDSEM_effects = c("phi(s)_21"),
  ranef_pred = "group_bi",
  inno_covs_zero = T)

#mlts_paths(model)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("dayphase_counter_0", "rnt"),
  iter = 1000
)

summary(fit)
mlts_standardized(fit)
```

* Try adding the momentary effect of the mindfulness intervention
```{r}
bolzen$intervene = ifelse(bolzen$timepoint != "t0", 1, 0)

bolzen %>% filter(!(id %in% c(36, 51,88))) %>%
  ggplot(aes(x = timepoint, y = rnt, group = id, color = as.factor(id))) +
  stat_smooth(method = "lm", se = F, alpha = 0.3) +
  theme(legend.position = "none") + 
  facet_grid(group~phase+time_window)


model = mlts_model(
  q = 3, 
  max_lag = 0, 
  is_exogenous = c(1,3),
  incl_rDSEM_effects = c("phi(s)_21"),
  incl_t0_effects = c("phi(0)_23"),
  fixef_zero = c("phi(1)_23"),
  ranef_pred = "group_bi",
  inno_covs_zero = T)

#mlts_paths(model)

fit = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(36, 51,88))),
  id = "id",
  ts = c("dayphase_counter_0", "rnt", "intervene"),
  iter = 1000
)

summary(fit)
mlts_standardized(fit)
```


## C: Moderation
```{r}
modVAR = mlts_model(
  q = 3,
  group = 2,
  fix_dynamics = T, 
  fix_inno_vars = T
  )

mlts_paths(modVAR)

# try to estimate: 
fit = mlts_fit(
  model = modVAR, 
  data = bolzen_spaced %>% filter(phase_bi == 1 & !(id %in% c(36,51,88))), 
  id = "id",
  group = "group",
  ts = c("negaff", "posaff", "rnt"),
  iter = 500,
  center_covs = F
  )

summary(fit, flag_signif = T)

mlts_plot(fit = fit, type = "fe", what = "Fixed effect")



```


### Umsetzungsvariable
```{r}
bolzen[,c("id", "time_window", "umsetzen")]

fit <- mlts_model(q = 1, max_lag = 0) %>%
  mlts_fit(model = ., data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(28,51))),
           id = "id", ts = "umsetzen",
           )
  
```


### TIME ON TASK TIME
```{r}
fit <- mlts_model(q = 1, max_lag = 0, ranef_pred = "group_bi") %>%
  mlts_fit(model = ., data = bolzen %>% filter(phase_bi == 1 & !(id %in% c(28,51))),
           id = "id", ts = "umsetzen",
           )

summary(lm(bolzen$audio_duration ~ bolzen$group))
summary(lm(bolzen$umsetzen ~ bolzen$group))
summary(lm(bolzen$pre_rtq_sum ~ bolzen$group))
summary(lm(bolzen$post_rtq_sum ~ bolzen$group))
summary(lm(bolzen$post_rtq_sum ~ scale(bolzen$pre_rtq_sum, scale=F) * bolzen$group))


```






# COMPLEX MODELS: 
```{r}
model = mlts_model(
  q = 3, 
  group = 2,
  is_exogenous = 1,
  fix_dynamics = T,
  fix_inno_vars = T,
  incl_interaction_effects = c(
    "phi(i)_2.1(1)2(1)",
    "phi(i)_2.1(1)3(1)",
    "phi(i)_3.1(1)2(1)",
    "phi(i)_3.1(1)3(1)"
  ))

mlts_paths(model)


# try to estimate: 
bolzen_spaced$inter = 0
bolzen_spaced$inter = ifelse(bolzen_spaced$timepoint == "t0", 1, bolzen_spaced$inter)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced %>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  group = "group",
  ts = c("inter", "negaff", "posaff"),
  iter = 500
  )

summary(fit)


```
# SIMPLIFIED VAR MODEL:

- POS and NEG affect with momentary link included as direct effect

  SEEMS TO WORK:
```{r}
model = mlts_model(
  q = 2, 
  fix_inno_vars = T,
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("negaff", "posaff"),
  iter = 500
  )

summary(fit)

```

* Stepwise build-up:
  - between-level covariate: group
```{r}
model = mlts_model(
  q = 2, 
  fix_inno_vars = T,
  ranef_pred = "group_bi",
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("negaff", "posaff"),
  iter = 500,
  center_covs = F
  )

summary(fit)

```
  

  - interaction: 
  VERY COMPLICATED ...
```{r}
model = mlts_model(
  q = 3, 
  fix_inno_vars = T,
  ranef_pred = "group_bi",
  fixef_zero = c("phi(1)_21", "phi(1)_12", "phi(1)_13", "phi(1)_23"),
  inno_covs_zero = T,
  
  incl_t0_effects = c("phi(0)_13", "phi(0)_23", "phi(0)_21"),
  incl_interaction_effects = c(
    "phi(i)_1.1(1)3(0)",
    "phi(i)_2.1(1)3(0)",
    "phi(i)_2.2(1)3(0)"
  ),
  is_exogenous = 3,
  )


mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("negaff", "posaff", "phase_bi"),
  iter = 500,
  center_covs = F
  )

summary(fit, digits = 2, flag_signif = T)

mlts_standardized(fit)

```

  - interaction via multiple-group model in intervention phase 
Very similar results, but: does not consider the changes between baseline and intervention
```{r}
model = mlts_model(
  q = 2, 
  group = 2,
  fix_inno_vars = T,
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced %>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("negaff", "posaff"),
  iter = 500,
  group = "group"
  )

summary(fit)

mlts_plot(fit)

```

  - interaction via multiple-group model in GROUPxPHASE 

```{r}
model = mlts_model(
  q = 2, 
  group = 4,
  fix_inno_vars = T,
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

bolzen_spaced$group_phase = paste0(bolzen_spaced$group, "_", bolzen_spaced$phase_bi)
bolzen_spaced$id_phase = paste0(bolzen_spaced$id, "_", bolzen_spaced$phase_bi)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced %>% filter(id != 30), 
  id = "id_phase",
  ts = c("negaff", "posaff"),
  iter = 500,
  group = "group_phase"
  )

summary(fit)

mlts_plot(fit, what = "Fixed effect")
mlts_plot(fit, what = "Fixed effect")
```




# SIMPLIFIED VAR MODEL:

- SAME AS ABOVE BUT POS AND RNT affect with momentary link included as direct effect

  SEEMS TO WORK:
```{r works}
model = mlts_model(
  q = 2, 
  fix_inno_vars = T,
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("rnt", "posaff"),
  iter = 2000
  )

summary(fit)

```

* Stepwise build-up:
  - between-level covariate: group
```{r works2}
model = mlts_model(
  q = 2, 
  fix_inno_vars = T,
  ranef_pred = "group_bi",
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("rnt", "posaff"),
  iter = 2000,
  center_covs = F
  )

summary(fit)

```
  
  - between-level covariate: group + prior differences in trait rnt
```{r}

model = mlts_model(
  q = 2, 
  fix_inno_vars = T,
  ranef_pred = c("group_bi","pre_rtq_sum_cen"),
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

summary(bolzen_spaced$pre_rtq_sum_cen)

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("rnt", "posaff"),
  iter = 2000,
  center_covs = F
  )

summary(fit)


bolzen_spaced$pre
```

  - interaction: 
  VERY COMPLICATED ...
```{r}
model = mlts_model(
  q = 3, 
  fix_inno_vars = T,
  ranef_pred = "group_bi",
  fixef_zero = c("phi(1)_21", "phi(1)_12", "phi(1)_13", "phi(1)_23"),
  inno_covs_zero = T,
  
  incl_t0_effects = c("phi(0)_13", "phi(0)_23", "phi(0)_21"),
  incl_interaction_effects = c(
    "phi(i)_1.1(1)3(0)",
    "phi(i)_2.1(1)3(0)",
    "phi(i)_2.2(1)3(0)"
  ),
  is_exogenous = 3,
  )


mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced, #%>% filter(phase_bi == 1 & !(id %in% c(NA))), 
  id = "id",
  ts = c("rnt", "posaff", "phase_bi"),
  iter = 2000,
  center_covs = F
  )

summary(fit, digits = 2, flag_signif = T)

mlts_standardized(fit)

```

  - interaction via multiple-group model in intervention phase 
Very similar results, but: does not consider the changes between baseline and intervention
```{r}
model = mlts_model(
  q = 2, 
  group = 2,
  fix_inno_vars = T,
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced %>% filter(phase_bi == 1 & !(id %in% c(36,51,88))), 
  id = "id",
  ts = c("rnt", "posaff"),
  iter = 2000,
  group = "group"
  )

summary(fit)

mlts_plot(fit)

```

  - interaction via multiple-group model in GROUPxPHASE 

```{r}
model = mlts_model(
  q = 2, 
  group = 4,
  fix_inno_vars = T,
  fixef_zero = c("phi(1)_21", "phi(1)_12"),
  incl_t0_effects = "phi(0)_21",
  inno_covs_zero = T
  )

mlts_paths(model)

bolzen_spaced$group_phase = paste0(bolzen_spaced$group, "_", bolzen_spaced$phase_bi)
bolzen_spaced$id_phase = paste0(bolzen_spaced$id, "_", bolzen_spaced$phase_bi)

fit = mlts_fit(
  model = model, 
  data = bolzen_spaced %>% filter(!(id %in% c(19,36,51,62,88))), 
  id = "id_phase",
  ts = c("rnt", "posaff"),
  iter = 2000,
  group = "group_phase"
  )

summary(fit)

mlts_plot(fit, what = "Fixed effect")
mlts_plot(fit, what = "Fixed effect")
```



