```{r, include=FALSE}
library(ggplot2)
library(mlts)
library(rstan)
```



# Live Session I (30 min)

* **Session Aims**:
  * Get familiar with `mlts_model()` syntax
  * Encounter common problems while fitting models
  * Check assumptions of AR models


* **What to do?** In groups of two (if you like),
  * fit a simple AR model for the construct *repetitive negative thinking* (variable `rnt` in dataset `bolzen_baseline`) or use your own data and interpret model results with your partner
  * If you work with the `bolzen` data, run the R script `bolzen_prepare.R` in the folder *live_session* once
    * Alternatively, just work in the R script `live_session1.R` directly and run the first 11 lines of code
  * Important: For practical reasons, in this exercise we only use the `bolzen_baseline` data frame for all analyses (without `tinterval`). Please do the same but keep in mind that the results may be taken with a large grain of salt.
  * Set `seed = 1234` in `mlts_fit()`, to replicate the exact results

---

## 1. AR(1) model for repetitive negative thinking {.scrollable-code}

::: {.panel-tabset}

### Task

Fit a model for repetitive negative thinking!

**Model:** 

- AR(1) model 

**Variables:** 

- Use repetitive negative thinking (`rnt`) as time series variable

**Data:**

- Use only data of the baseline phase (dataset `bolzen_baseline`)

**To explore:**

- You'll likely get an error message before fitting the model - try to understand and fix it!
- When you get `mlts_fit()` to work, only fit the model with 500 MCMC iterations
- What do you realize after fitting? (Go to next task after taking note)

### Solution

::: {.output-scroll}

```{r}
#| eval: true
#| echo: false

mlts_fit1 <- readRDS("./live_session/live_session1_fits/mlts_fit1.rds")
```


```{r}
#| eval: false
#| echo: true

# model in mlts
mlts_m1 <- mlts_model(q = 1)
mlts_m1
# mlts_model_paths(mlts_m1)

# fit
mlts_fit1 <- mlts_fit(
  model = mlts_m1,
  data = bolzen_baseline,
  id = "id",
  ts = "rnt",
  iter = 2000
) # throws an error


# Error Message:
# "Within-cluster variance is zero for indicator rnt in cluster 19"
# - means that there is no variation in the variable for that id
#   (i.e., always answered in the same category)

# Solution: filter out id 19 manually:

bolzen_clean <- bolzen_baseline %>%
  filter(id != 19)

# refit
mlts_fit1 <- mlts_fit(
  model = mlts_m1,
  data = bolzen_clean,
  id = "id",
  ts = "rnt",
  iter = 2000
) # throws another error

# better solution: filter out all ids with Within-cluster variance == zero
# on the respective indicator
# can be achieved, e.g., with n_distinct from dplyr
bolzen_clean <- bolzen_baseline %>%
  group_by(id) %>%
  filter(n_distinct(rnt, na.rm = TRUE) > 1)

# refit
mlts_fit1 <- mlts_fit(
  model = mlts_m1,
  data = bolzen_clean,
  id = "id",
  ts = "rnt",
  iter = 500
) # this works
```


```{r}
#| eval: true
#| echo: true

# summary
summary(mlts_fit1) # mixing is bad (see results section)

# inspect traceplots
traceplot(mlts_fit1$stanfit)
```


:::

:::


---

## 2. AR(1) model for repetitive negative thinking with restriction {.scrollable-code}

::: {.panel-tabset}

### Task

Without further restrictions, the model from exercise 1 does not converge.
Inspect the parameters from the fitted model again and try to find restriction(s) such that the model converges!

**Model:** 

- AR(1) model with some parameters restricted

**Variables:** 

- Use repetitive negative thinking (`rnt`) as time series variable

**Data:**

- Use only data of the baseline phase (dataset `bolzen_baseline`)

**To explore:**

* Inspect traceplots

### Solution

::: {.output-scroll}

```{r}
#| eval: true
#| echo: false

mlts_fit2 <- readRDS("./live_session/live_session1_fits/mlts_fit2.rds")
```


```{r}
#| eval: false
#| echo: true

# model in mlts
mlts_m2 <- mlts_model(q = 1, fix_inno_vars = TRUE)
mlts_m2
# mlts_model_paths(mlts_m1)

# fit
mlts_fit2 <- mlts_fit(
  model = mlts_m2,
  data = bolzen_clean,
  id = "id",
  ts = "rnt",
  iter = 2000
)
```


```{r}
#| eval: true
#| echo: true

# summary
summary(mlts_fit2) # mixing is bad (see results section)

# inspect traceplots
traceplot(mlts_fit2$stanfit)
```


:::

:::

---

## 3. Check assumptions {.scrollable-code}

::: {.panel-tabset}

### Task

(Usually before fitting models,) We need to check the assumptions of the AR model.

**Stationarity**

- Check for stationarity using graphical checks or other methods, if you have more ideas!

**Variables:** 

- Use repetitive negative thinking (`rnt`) as time series variable

**Data:**

- Use only data of the baseline phase (dataset `bolzen_baseline`)

**To explore:**

* *Advanced.* If you have time left, check the script `bolzen_prepare.R` and try to understand the steps involved in creating a time variable (called `int_time_spaced`) that takes the gaps in measurement into account!

### Solution

::: {.output-scroll}


```{r}
#| eval: true
#| echo: true

# plot trend
ggplot(aes(x = obs_counter, y = rnt),
       data = bolzen_baseline) +
  geom_smooth(aes(group = id),
              method = "lm", se = FALSE,
              color = "grey", linewidth = .5) +
  geom_smooth(method = "lm") +
  theme_minimal(base_size = 14) +
  labs(y = "RNT", x = "(Integer) Time Point")

# other method: fit growth curve model in lme4
library(lme4)

# fit growth curve model
gcm <- lmer(rnt ~ 1 + obs_counter + (1 + obs_counter | id),
            data = bolzen_baseline)

# check summary
summary(gcm) # literally no growth
```


:::

:::

