
# Model Extensions 

```{r}
dir.create("sections/05-model_extensions/")

# load all fitted models
for(i in Sys.glob("sections/05-model_extensions/*.Rdata")){
  load(i)
}
```

---

### Vector Autoregressive Model

- Allows estimation of time-lagged effects between multiple variables
  - Autoregressive and cross-lagged (CL) effects
- In multivariate models, the number of random effects can quickly increase
  - Estimation times can quickly increase, especially for large *N*
  - Yielding acceptable model convergence criteria can be much challenging
  
- We will illustrate this for the positive and negative affect composite scores

---

:::: {.columns}

::: {.column width="35%"}

#### Decomposition

Applied to each variable:
$$
\begin{matrix}
Y_{1it}^{w} &= Y_{1it} - \mu_{1i} \\
Y_{2it}^{w} &= Y_{2it} - \mu_{2i} \\
\end{matrix}
$$ 

#### Within-level
$$
\begin{matrix}
Y_{1it}^w &= \phi_{11i} Y_{1i(t-1)}^{w} + \phi_{12i} Y_{2i(t-1)}^{w} + \zeta_{1it} \\
Y_{2it}^w &= \phi_{21i} Y_{1i(t-1)}^{w} + \phi_{22i} Y_{2i(t-1)}^{w} + \zeta_{2it} \\
\end{matrix}
$$

$$
\begin{matrix}
\text{with}\ \zeta_{1it} \sim N(0, \sigma_{\zeta 1i}^2) \\
\text{with}\ \zeta_{2it} \sim N(0, \sigma_{\zeta 2i}^2)
\end{matrix}
$$ 

#### Between-level
$$
\begin{matrix}
\mu_{1i}  = \gamma_{\mu1}  + \upsilon_{\mu1i} \\
\mu_{2i}  = \gamma_{\mu1}  + \upsilon_{\mu1i} \\
\phi_{11i} = \gamma_{\phi11} + \upsilon_{\phi 11i} \\
\phi_{12i} = \gamma_{\phi12} + \upsilon_{\phi 12i} \\
[...] \\
ln(\sigma_{\zeta 2i}^2) = \gamma_{ln(\sigma^2_{\zeta2})} + \upsilon_{ln(\sigma^2_{\zeta 2i})}
\end{matrix}
$$ 

$$
\text{with}\ \mathbf{\upsilon} \sim MVN(\mathbf{0},\mathbf{\Sigma})
$$ 

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

#### Model Specification

```{r, echo = T}
model = mlts_model(
  q = 2,                 # number of time-varying constructs
  inno_covs_zero = TRUE  # assume independent innovations (r = 0)
  )
```


```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.8, scale_within_inno = 0.22,
           cex_title = 0.8, scale_between = 0.32, asp_w_b = 0.4)
```

:::
::::

--- 

### Fixed Effect Parameters in the `model`-object
```{r}
#| echo: true
model[1:8,c("Type","Param","Param_Label")]
```

- `phi(1)_11` and `phi(1)_22` are the AR effects of variable 1 and 2
- `phi(1)_21` and `phi(1)_12` are the first-order CL effect between variables 1 and 2. The first number indicates the DV (outcome) and the second number the IV (predictor) 

- The order of the variables (1 or 2) will be determined by the order in which the variable names are passed to the `ts`-argument in `mlts_fit()`. 

--- 

## Model Estimation | A Detailled Workflow

::: {.column width="55%"}

- For ESM datasets of moderate sample size (N ~ 100 & T ~ 70), estimating VAR(1) models with random effects for all time-lagged effects and innovation variances can be difficult
  - Insufficient model convergence indices & long estimation times 

- Usually the result of model misfit 
  - Known issue: Parameter estimates close to boundary values
    - Low random effect variances
    - Random effect correlations close to 1 (or -1)
  - Preliminary estimation runs (with ~ 500-1000 iterations) can help to diagnose model misfit at an early stage 

:::

. . .

::: {.column width="5%"}

:::

::: {.column width="40%"}
```{r}
#| eval: false
#| echo: true
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 500)
```
:::

---

### Preliminary Run | Results

:::: {.columns}

::: {.column width="45%"}

```{r, eval = F}
fit_VAR = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR, file = "05-model_extensions/1.1_VAR_500.Rdata")

fit_VAR_l = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  iter = 2000,
  seed = 1234)

save(fit_VAR_l, file = "05-model_extensions/1.1_VAR_2000.Rdata")

```

```{r}
load("sections/05-model_extensions/1.1_VAR_500.Rdata")
load("sections/05-model_extensions/1.1_VAR_2000.Rdata")
```

```{r}
#| eval: false
#| echo: true
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 500)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time: Longest chain took 14.9 minutes


:::
::::

---

### Preliminary Run | Results for `iter = 2000`

:::: {.columns}

::: {.column width="45%"}

```{r}
#| eval: false
#| echo: true
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR_l = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR_l, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time: Longest chain took 36.5 minutes

:::
::::

--- 

### Running Model With Constraints

```{r, eval = F}
model_r = mlts_model(q = 2, inno_covs_zero = T, 
                     ranef_zero = "phi(1)_21")

fit_VAR_r = mlts_fit(
  model = model_r,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_r, file = "05-model_extensions/1.1_VAR_r1_500.Rdata")

fit_VAR_r_l = mlts_fit(
  model = model_r,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_r_l, file = "05-model_extensions/1.1_VAR_r1_2000.Rdata")

model_r2 = mlts_model(q = 2, inno_covs_zero = T, 
                     ranef_zero = c("phi(1)_21", "phi(1)_12"))

fit_VAR_r2 = mlts_fit(
  model = model_r2,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_r2, file = "05-model_extensions/1.1_VAR_r2_500.Rdata")

fit_VAR_r2_l = mlts_fit(
  model = model_r2,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_r2_l, file = "05-model_extensions/1.1_VAR_r2_2000.Rdata")
```


::: {.column width="25%"}

:::

::: {.column width="7%"}
**iter**
:::

::: {.column width="7%"}
**Time**
:::

::: {.column width="20%"}
**Diagnostics**
:::

::: {.column width="41%"}
**Crit. Parameters**
:::


```r
mlts_model(q = 2, inno_covs_zero = T)
```

::: {.column width="25%"}

:::

::: {.column width="7%"}
500
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun <- function(fit){
  max.rhat = round(max(fit$pop.pars.summary$Rhat),3)
  min.ess_bulk = min(fit$pop.pars.summary$Bulk_ESS)
  min.ess_tail = min(fit$pop.pars.summary$Tail_ESS)
  n_div  = rstan::get_num_divergent(fit$stanfit)
  cat(paste0(
    "Max Rhat     = ", max.rhat, "\n",
 #   "Min Tail ESS = ", min.ess_tail, "\n",
#    "Min Bulk ESS = ", min.ess_bulk, "\n",
    "N divergent  = ", n_div
  ))
}
diag_fun(fit_VAR)
```
:::

::: {.column width="41%"}
```{r}
fun_crit <- function(fit){
  cols = c("Param", "Rhat","Bulk_ESS", "Tail_ESS")
  a = which(fit$pop.pars.summary$Tail_ESS == min(fit$pop.pars.summary$Tail_ESS))
  b = which(fit$pop.pars.summary$Bulk_ESS == min(fit$pop.pars.summary$Bulk_ESS))
  c = which(fit$pop.pars.summary$Rhat == max(fit$pop.pars.summary$Rhat))
  fit$pop.pars.summary[unique(c(a,b,c)), cols]  
}

fun_crit(fit_VAR)
```
:::

<!-- new row -->

::: {.column width="25%"}

:::

::: {.column width="7%"}
2000
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_l$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_l)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_l)
```
:::

<!-- new row -->

:::{.vspace}
:::

```r
mlts_model(q = 2, inno_covs_zero = T, ranef_zero = "phi(1)_21")
```


::: {.column width="25%"}

:::

::: {.column width="7%"}
500
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r)
```
:::

<!-- new row -->

::: {.column width="25%"}

:::

::: {.column width="7%"}
2000
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r_l$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r_l)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r_l)
```
:::

<!-- new row -->


:::{.vspace}
:::

```r
mlts_model(q = 2, inno_covs_zero = T, ranef_zero = c("phi(1)_21","phi(1)_12"))
```


::: {.column width="25%"}

:::

::: {.column width="7%"}
500
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r2$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r2)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r2)
```
:::

<!-- new row -->

::: {.column width="25%"}

:::

::: {.column width="7%"}
2000
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r2_l$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r2_l)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r2_l)
```
:::

---

### Take-Away

- Estimation times can quickly increase, especially when the model does not fit the data
  - here: random effect variances of the CL effects 
- Preliminary test runs with less iterations can usually detect misfit early on 
  - Strong recommended especially for complex models 
- When the model fits the data, running more iterations is usually less costly 


---

### Take-Away

- Estimation times can quickly increase, especially when the model does not fit the data
  - here: random effect variances of the CL effects 
- Preliminary test runs with less iterations can usually detect misfit early on 
  - Strong recommended especially for complex models 
- When the model fits the data, running more iterations is usually less costly 


:::{.vspace}
:::

:::{.large}
**Are we Done here?**
:::


---

### Are we Done here?

- Let's recap, why did the unrestricted VAR(1) model fail?
- Any suggestions what might change the results?

```{r}
summary_lines <- capture.output(summary(fit_VAR_r2_l, digits = 2))
cat("[...]")
cat(summary_lines[11:21], sep = "\n")  # e.g. first 10 
cat("[...]")
```

---

### Refitting the Model with `tinterval`

```{r, eval = F}
fit_VAR_tint = mlts_fit(
  model = model,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_tint, file = "05-model_extensions/1.1_VAR_tint_500.Rdata")

fit_VAR_tint_l = mlts_fit(
  model = model,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_tint_l, file = "05-model_extensions/1.1_VAR_tint_2000.Rdata")


fit_VAR_tint_r = mlts_fit(
  model = model_r,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_tint_r, file = "05-model_extensions/1.1_VAR_tint_r_500.Rdata")

fit_VAR_tint_r_l = mlts_fit(
  model = model_r,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_tint_r_l, file = "05-model_extensions/1.1_VAR_tint_r_2000.Rdata")

```

:::: {.columns}

::: {.column width="45%"}

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: 8-10
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR_tint_l = mlts_fit(
  model = model,
  data = bolzen_grid,
  id = "id",
  ts = c("posaff", "negaff"),
  tinterval = 1,
  time = "int_time_spaced",
  max_NA_seq = 4,
  seed = 1234,
  iter = 2000)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR_tint_l, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time:
```{r, eval = F}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_tint_l$stanfit),1,sum))/60,2))
```
Longest chain took 16.1 minutes


:::
::::

---

### Refitting the Model with `tinterval` II

:::: {.columns}

::: {.column width="45%"}

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: 1-13|2
model_r = mlts_model(q = 2, inno_covs_zero = T, 
                     ranef_zero = "phi(1)_21")

fit_VAR_tint_r_l = mlts_fit(
  model = model_r,
  data = bolzen_grid,
  id = "id",
  ts = c("posaff", "negaff"),
  tinterval = 1,
  time = "int_time_spaced",
  max_NA_seq = 4,
  seed = 1234,
  iter = 2000)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR_tint_r_l, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time:
```{r, eval = F}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_tint_r_l$stanfit),1,sum))/60,2))
```
Longest chain took 15.1 minutes


:::
::::



---



### Innovation Covariance - Three modeling options

:::: {.columns}

::: {.column width="30%"}

1. Independent: 
```{r}
#| code-line-numbers: 3-3
#| eval: false
#| echo: true
mlts_model(
  q = 2,
  inno_covs_zero = TRUE
  )
```


:::

::: {.column width="5%"}
:::

::: {.column width="30%"}

2. Constant innovation correlation 
```{r}
#| code-line-numbers: 3-4
#| eval: false
#| echo: true
mlts_model(
  q = 2,
  inno_covs_zero = FALSE,  # default
  fix_inno_covs = TRUE     # default
  )
```

:::

::: {.column width="5%"}

:::

::: {.column width="30%"}

3. Random Innovation covariance 
```{r}
#| code-line-numbers: 3-5
#| eval: false
#| echo: true
mlts_model(
  q = 2,
  inno_covs_zero = FALSE,
  fix_inno_covs = FALSE,
  inno_covs_dir = "pos" # or "neg"
  )
```

:::
::::

:::: {.columns}

::: {.column width="30%"}
Implies:

$$
\begin{bmatrix}
\zeta_{1it} \\
\zeta_{2it} 
\end{bmatrix} 
\sim MVN 
\left(
\begin{bmatrix}
0 \\ 0
\end{bmatrix},
\begin{bmatrix}
\sigma_{\zeta 1i}^2 & 0 \\ 
0 & \sigma_{\zeta 2i}^2
\end{bmatrix}
\right)
$$ 
:::

::: {.column width="5%"}

:::

::: {.column width="30%"}
Implies:

$$
\begin{bmatrix}
\zeta_{1it} \\
\zeta_{2it} 
\end{bmatrix} 
\sim MVN 
\left(
\begin{bmatrix}
0 \\ 0
\end{bmatrix},
\begin{bmatrix}
\sigma_{\zeta 1i}^2 & 0 \\ 
\sigma_{\zeta 12i}^2 & \sigma_{\zeta 2i}^2
\end{bmatrix}
\right)
$$ 
where
$$
\sigma_{\zeta 12i}^2 = \rho_{\zeta12} * \sigma_{\zeta 1i} * \sigma_{\zeta 2i}
$$
$\rho_{\zeta12}$ without index $i$ --> constant across clusters

:::

::: {.column width="5%"}

:::

::: {.column width="30%"}
Implies:
$$
\begin{matrix}
\zeta_{1it} = 1 * \eta_{\zeta,it} + \delta_{1it} \\
\zeta_{2it} = 1 * \eta_{\zeta,it} + \delta_{2it} \\
\end{matrix} 
$$

$$
\begin{matrix}
\delta_{1it} \sim N(0, \sigma_{\delta 1i}^2)\\
\delta_{2it} \sim N(0, \sigma_{\delta 2i}^2)\\
\eta_{\zeta,it} \sim N(0, \sigma_{\eta \zeta i}^2)
\end{matrix} 
$$ 

:::
::::

---


### Between-level Covariates  

Person-level variables as predictor of individual differences in any of the dynamic model parameters:  

:::: {.columns}

::: {.column width="35%"}

#### Decomposition

- Latent Person-mean-centering:

$$
Y_{it}^{w} = Y_{it} - \mu_{i}
$$ 

#### Within-level
$$
Y_{it}^w = \phi_i Y_{i(t-1)}^{w} + \zeta_{it} \\
\text{with}\ \zeta_{it} \sim N(0, \sigma_{\zeta i}^2)
$$ 


#### Between-level
$$
\begin{matrix}
\mu_i  &=& \gamma_{\mu}  &+& b_{\mu}Group_i  &+& \upsilon_{\mu i} \\
\phi_i &=& \gamma_{\phi} &+& b_{\phi}Group_i  &+& \upsilon_{\phi i} \\
ln(\sigma_{\zeta i}^2) &=& \gamma_{ln(\sigma^2_\zeta)} &+& b_{ln(\sigma_\zeta^2)}Group_i  &+& \upsilon_{ln(\sigma^2_{\zeta i})}
\end{matrix}
$$ 

$$
\text{with}\ \mathbf{\upsilon} \sim MVN(\mathbf{0},\mathbf{\Sigma})
$$ 

:::

::: {.column width="10%"}
<!-- Add some space between columns -->
:::

::: {.column width="55%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 1, 
  ranef_pred = "group_bi"  # name of the variable in the dataset
  )
```

We use the dummy-coded treatment condition variable (i.e., `group_bi`).

```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3
#| fig-dpi: 1000
#| fig-align: left


mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.25,
           asp_w_b = 0.6,
           scale_within_inno = 0.3,
           cex_title = 0.8)
```

:::
::::

---

### Important

  - By default between-level covariates will be grand-mean centered to allow a clearer interpretation of the dynamic model parameters 
  - For binary predictor variables, you can turn off this behavior by setting ` mlts_fit(..., center_covs = FALSE)`

---

### Prediction of Time-Stable Outcomes

Person-level variables as outcomes to be predicted by person-specific parameters: 

:::: {.columns}

::: {.column width="55%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 1, 
  out_pred = "Out" # name of the variable in the dataset
  )
```

```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left


mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.2,
           asp_w_b = 0.6,
           scale_within_inno = 0.3,
           cex_title = 0.8)
```
:::

::: {.column width="10%"}
<!-- Add some space between columns -->
:::

::: {.column width="35%"}
**Further options**

- For multiple outcomes: 
  - `out_pred = c("Out1", "Out2")`
- Specific paths: 
  - `list(c("Out" = "mu_1"))`
  - Uses only individual mean levels (`mu_1`) as predictor
- Include observed binary or continuous person-level variables: 
  - `out_pred_add_btw = "Z"`, or:
  - `list(c("Out" = "mu_1", "Z"))`
  - where `"Z"` refers to a variable in the dataset


:::

::::

---


### Multiple-Group Models

:::{.large}

- Simultaneous estimation of the *same dynamic model* *separately for a specified number of groups* 

- As shown before, differences in model parameters can be estimated by adding the group-variable as predictor of individual-specific effects. 
  - **But**: only possible for random effects.

- Multiple-group models: Allow estimation of group-specific model parameters 
  - Differences in average (i.e. fixed) effects, random effect variance, random effect correlation, 
  innovation correlations

- Possible scenarios where this is relevant:
  - Relatively low number of measurements which would require estimating some model parameters as constant but group-specific differences therein are of interest

:::  

---

### Multiple-Group Models

::: {.column width="50%"}
#### Model Specification 
```{r}
#| echo: true
VAR1_g <- mlts_model(
  q = 2,
  group = 2,  
  fix_dynamics = TRUE,
  fix_inno_vars = TRUE,
  )
```

#### Model Estimation
```{r}
#| echo: true
#| eval: false
fit <- mlts_fit(
  model = VAR1_g,
  data = bolzen %>% filter(phase == "i"),
  id = "id",
  ts = c("negaff","posaff"),
  group = "group",
  iter = 2000
)
```
```{r}
#save(fit, file = "05-model_extensions/4.1_MG.Rdata")
load("sections/05-model_extensions/4.1_MG.Rdata")
```
:::

::: {.column width="50%"}
#### Results
```{r}
summary_lines <- capture.output(summary(fit, digits = 2))
cat(summary_lines[1:13], sep = "\n")  
```
:::

---

### Multiple-Group Models

#### Results
::: {.column width="50%"}

```{r}
cat(summary_lines[14:35], sep = "\n")  
```

:::

::: {.column width="50%"}
```{r}
cat(summary_lines[37:58], sep = "\n")  
```
:::


---

### Multiple-Group Models

#### Group-Comparisons

- Based on the group-specific estimates, we can run comparison tests
- The procedure to conduct these tests is the same for any model parameter (also random effect variances or random effect correlations)
- For example, are there group-differences in the average CL effect of `poasaff` at $t-1$ on `negaff` at $t$ (i.e., `phi(1)_12`).

- We first extract the posterior draws of the group-specific estimates:
```{r}
#|echo: true
CL_dm = fit$posteriors[,,"dm: phi(1)_12"]
CL_gi = fit$posteriors[,,"gi: phi(1)_12"]
```

- Then, calculate the differences between the group-specific estimates for each draw:
```{r}
#|echo: true
Diff = as.vector( CL_dm - CL_gi )
```

- Now, we can calculate summary statistics: 
  - such as the difference parameters' posterior mean, 
```{r}
#|echo: true
mean(Diff)
```
  - or a 95% CI to test potential differences with $\alpha = .05$:
```{r}
#|echo: true
quantile(Diff, probs = c(0.025, 0.975))
```

---

### Interaction Effects on the Dynamic Within-Level


- Thus far, we assumed *AR*- or *CL*-effects to be invariant over time

- This assumption can be relaxed or tested, e.g., using **time-varying effects models** (e.g., Bringmann et al., 2017) 
  - gradual changes in intercepts, or time-lagged effects over time can be  

- Conceptually different: **Moderated VAR models** (Holtmann & Koslowski, 2025) as a multilevel extension of single-subject fixed moderated time series analysis (Adolf et al., 2017)
  
  - Moderation effects on the dynamic within-level where the moderator is known 
  - Inclusion of time-varying variables as moderators of dynamic regression effects (i.e., *AR* and *CL*)

**Possible research questions**

- Does engaging in emotion regulation strategies (e.g., rumination) at time $t-1$ impact the carry-over of negative affect at $t-1$ to the subsequent time point (i.e., moderated *AR*)? 



:::{.references}
Adolf, J. K., Voelkle, M. C., Brose, A., & Schmiedek, F. (2017). Capturing Context-Related Change in Emotional Dynamics via Fixed Moderated Time Series Analysis. *Multivariate Behavioral Research*, *52(4)*, 499–531. <https://doi.org/10.1080/00273171.2017.1321978>

Bringmann, L. F., Hamaker, E. L., Vigo, D. E., Aubert, A., Borsboom, D., & Tuerlinckx, F. (2017). Changing dynamics: Time-varying autoregressive models using generalized additive modeling. *Psychological Methods*, *22(3)*, 409–425. <https://doi.org/10.1037/met0000085>

Holtmann, J., & Koslowski, K. (2025). Modeling within-level latent interaction effects in multilevel vector-autoregressive models. *Behavior Research Methods*, *57(10)*, 277. <https://doi.org/10.3758/s13428-025-02694-3>

:::

---

### A Simple Example

**Do positive affect dynamics change from the baseline to the intervention phase?** 

:::{.vspace}
:::

:::: {.columns}

::: {.column width="45%"}

- For simplicity, we assume a time-varying binary indicator variable for the study phases (i.e. `phase_bi`)

- We want `phase_bi` to be treated as an exogenous variable:
  (1) it only serves as predictor
  (2) without latent-mean centering 


:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

#### Decomposition

- Latent Person-mean-centering only for positive affect ($PA$)

$$
PA_{it}^{w} = PA_{it} - \mu_{PA}
$$ 

#### Within-level
$$
PA_{it}^w = \phi_{21i} Phase_{it} + \phi_{22i} PA_{i(t-1)}^{w} + \phi_{2.21i} PA_{i(t-1)}^{w}*Phase_{it} + \zeta_{it} \\
$$

Implies:
$$
\begin{matrix}
\widehat{PA_{it}^w} &=& \phi_{22i} PA_{i(t-1)}^{w}  & \text{if}\ Phase_{it} = 0 \\
\widehat{PA_{it}^w} &=& \phi_{21i} + (\phi_{22i}+\phi_{2.21i})PA_{i(t-1)}^{w} & \text{if}\ Phase_{it} = 1\\
\end{matrix}
$$

$$
\text{with}\ \zeta_{it} \sim N(0, \sigma_{\zeta i}^2)
$$ 


:::

::::

--- 

### We build the model in a step-wise manner

::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  # two constructs: 1 = phase_bi, 2 = posaff
  q = 2
  )
```

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::


--- 

### We build the model in a step-wise manner

::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  # two constructs: 1 = phase_bi, 2 = posaff
  q = 2,                          
  # treat first construct as exogenous:
  is_exogenous = 1
  )
```

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::

--- 

### We build the model in a step-wise manner


::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 2,                          
  is_exogenous = 1,               
  
  # PA (t) regressed on phase_bi (t):
  incl_t0_effects = "phi(0)_21",  

  # additional constraints: 
  ## remove default first-order CL effect
  fixef_zero = "phi(1)_21",       
  ## for technical reasons, set
  inno_covs_zero = TRUE           
  )
```

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::

--- 

### We build the model in a step-wise manner

::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 2,                          
  is_exogenous = 1,               
  incl_t0_effects = "phi(0)_21",  
  fixef_zero = "phi(1)_21",       
  inno_covs_zero = TRUE,            
  
  # include interaction effect:
  incl_interaction_effects = "phi(i)_2.2(1)1(0)" 
  )
```


#### Breakdown of `phi(i)_2.2(1)1(0)`

$$
\begin{array}{lll}
\text{phi(i)_} &\longrightarrow& \text{interaction effect} \\
\text{2.}      &\longrightarrow& \text{DV (PA)} \\
2(1)           &\longrightarrow& \text{IV I (PA) at lag 1} \\ 
1(0)           &\longrightarrow& \text{IV II (SP) at lag 0} \\
\end{array}
$$

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::

---

### Multiple-Indicator Factor Models

- DSEM model are oftentimes run for aggregate scores 
  - Composite across multiple ESM items 
  - For example, *happy*, *relaxed*, and *cheerful* as indicators of positive affect
  - Measurement error in observed variables is not accounted for

- Consequences of disregarding measurement error
  - For time series variables, detrimental effects on the reliability of (individual) lagged effect estimates [e.g., @duReliabilitiesIntraindividualVariability2018]   
  - Underestimated AR effects in AR models [e.g., @schuurmanIncorporatingMeasurementError2015]
  - Potentially distorted estimates of CL relations in VAR models [e.g., @schuurmanMeasurementErrorPersonspecific2019]
  - Low reliability of person-specific estimates can impact their use as predictors 

- Solution:  
  - including a measurement model (both at the between- and/or the within-level) and
  - modeling the within-person dynamics for the underlying latent factors
  - requires that multiple parallel indicators of the same construct are available

---

:::: {.columns}

::: {.column width="35%"}

#### Decomposition
$$
\begin{matrix}
Y_{1.1it}^{w} &= Y_{1.1it} - \mu_{1.1i} \\
[. . .] \\
Y_{q.pit}^{w} &= Y_{q.pit} - \mu_{q.pi} \\
\end{matrix}
$$ 

#### Within-level measurement model
$$
\begin{matrix}
Y_{q.1it}^w &= \lambda_{q.p}^w \eta_{qit}^{w} + \varepsilon_{qit}^w \\
\end{matrix} \\
\text{with}\ \varepsilon_{qit}^w \sim N(0, \sigma_{\varepsilon_w,q.p}^2)
$$
For model identification: $\lambda_{q.1}^w = 1$.

#### Dynamic model (here AR(1))
$$
\begin{matrix}
\eta_{qit}^{w} &= \phi_{11i} \eta_{qi(t-1)}^{w} + \zeta_{qit} \\
\end{matrix} \\
\text{with}\ \zeta_{1it} \sim N(0, \sigma_{\zeta 1i}^2) \\
$$

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

#### Model Specification

```{r, echo = T}
model = mlts_model(
  q = 1,              # number of time-varying constructs
  p = 3,              # number of manifest indicator
  btw_factor = FALSE, # no common between-level factor 
  )
```


```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 4, width = 6, cex_b = 0.8, 
           asp_decomp = 0.4,
           scale_within_inno = 0.22,
           cex_title = 0.8, scale_between = 0.38, asp_w_b = 0.4)
```
:::
::::


---

:::: {.columns}

::: {.column width="35%"}

#### Decomposition
$$
\begin{matrix}
Y_{1.1it}^{w} &= Y_{1.1it} - \mu_{1.1i} \\
[. . .] \\
Y_{q.pit}^{w} &= Y_{q.pit} - \mu_{q.pi} \\
\end{matrix}
$$ 


#### Between-level (indicator-specific person means)
$$
\begin{matrix}
\mu_{1.1i}  = \gamma_{\mu1.1}  + \upsilon_{\mu1.1i} \\
[...] \\
\mu_{q.pi}  = \gamma_{\mu q.p}  + \upsilon_{\mu q.pi} \\
\phi_{11i} = \gamma_{\phi11} + \upsilon_{\phi 11i} \\
ln(\sigma_{\zeta 2i}^2) = \gamma_{ln(\sigma^2_{\zeta2})} + \upsilon_{ln(\sigma^2_{\zeta 2i})}
\end{matrix} \\
\text{with}\ \mathbf{\upsilon} \sim MVN(\mathbf{0},\mathbf{\Sigma})
$$ 
:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

#### Model Specification

```{r, echo = T}
model = mlts_model(
  q = 1,              # number of time-varying constructs
  p = 3,              # number of manifest indicator
  btw_factor = FALSE, # no common between-level factor 
  )
```


```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 4, width = 6, cex_b = 0.8, 
           asp_decomp = 0.4,
           scale_within_inno = 0.22,
           cex_title = 0.8, scale_between = 0.38, asp_w_b = 0.4)
```
:::
::::

---

### Extended modeling options

::: {.column width="50%"}

#### Common betwen-level factor 

```{r}
#| echo: true
model = mlts_model(
  q = 1, 
  p = 3, 
  btw_factor = TRUE)
```

```{r}
#| fig-width: 5
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.4, cex_b = 0.9)
```

:::

  
::: {.column width="50%"}

#### Complex VAR(1) structure 

```{r}
#| echo: true
model = mlts_model(
  q = 2, 
  p = c(3,4), 
  btw_model = list(1:3, NA),
  fix_dynamics = T,
  fix_inno_vars = T)
```

```{r}
#| fig-width: 7
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.6,
           cex_w = 0.7)
```

:::  
  
---  
  
### Extended modeling options 
  
::: {.column width="50%"}

#### Assuming cross-level measurement invariance 

```{r}
#| echo: true
model = mlts_model(
  q = 1, 
  p = 3, 
  btw_factor = TRUE, 
  equal_loads_levels = T
  )
```

```{r}
#| fig-width: 5
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.4, cex_b = 0.9)
```

:::
  
  
::: {.column width="50%"}

#### Additional constraints

```{r}
#| echo: true
model = mlts_model(q = 1, p = 3)

# Fix loading parameter to 1
model$Constraint[model$Param == "lambdaW_1.2"] <- "= 1"
# Fix error variance to be 0
model$Constraint[model$Param == "sigmaB_1.2"] <- "= 0"
# Remove constraints (agian)
model$Constraint[model$Param == "sigmaB_1.2"] <- "free"

```

```{r}
#| fig-width: 5
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.45, cex_b = 0.9)
```

:::    

--- 


### Censoring Models

- Approach to modeling ESM data with floor or ceiling effects
- Floor effects challenge the normality assumptions underlying standard DSEM models 
- Specific assumption about the data generating process underlying floor effects 
  - Measurement-related censoring (i.e., observations are censored during measurement)
  - Implies: There is substantial variation in the underlying construct not captured by the item that exhibits floor effects
  
#### Evidence in favor of the censoring notion?

::: {.column width="60%"}
```{r}
#| fig-width: 12
#| fig-height: 5
#| fig-pos: left

# describe_ts(data = bolzen, ts = c("nervoes", "entspannt"), group = "id") %>%
#   arrange(desc(floor_pc))

bolzen %>%
  filter(id %in% c(11, 35)) %>%
  group_by(id) %>%
  mutate(
    entspannt_r = 8-entspannt
  ) %>%
  
  ggplot(aes(y = nervoes, x = obs_counter)) +
  geom_line() +
  geom_line(aes(y = entspannt_r), color = "darkgreen") +
  geom_point(alpha = 0.5) +
  geom_point(aes(y = entspannt_r), color = "darkgreen", alpha = 0.5) +
  geom_text(x = 87, y = 1.2, hjust = 0, label = "not nervous") +
  geom_text(x = 87, y = 1.9, hjust = 0, label = "relaxed",color = "darkgreen") +
  geom_text(x = 87, y = 6.8, hjust = 0, label = "nervous") +
  geom_text(x = 87, y = 6.1, hjust = 0, label = "not relaxed",color = "darkgreen") +
  scale_x_continuous(limits = c(0,95), expand = c(0,1)) +
  scale_y_continuous(limits = c(1,7), breaks = 1:7) +
  labs(x = "# Beep", y = NULL) +
  facet_grid(rows = vars(id)) +
  theme_classic() +
  theme(
    panel.spacing.y = unit(0.5, "cm")
  )


```
:::

::: {.column width="40%"}

- Inspecting unipolar items capturing the opposing ends of a continuum can be informative
- Here, as illustrated for two individuals, when subjects report to feel not nervous at all, they still show substantial variation in the relaxed item in these moments
- Hence, on the level of the broader construct, this suggests at qualitative differences in states of tense arousal beyond what is captured by the nervous item


:::


--- 

::: {.column width="80%"}

#### Impact of Ignoring Censoring-Related Floor Effects 

- Introduces spurious positive dependencies between person-specific estimates 
  - mean-variance-relation(s) most affected
- Censoring on the level of individual ESM items can affect estimates of analyses that use composite scores across indicators that exhibit floor effects
  - Composite scores often show more desirable distributional properties (e.g., less floor effects)
  - However, adverse effects on parameter estimates are not fully resolved by item aggregation
  - → Censored multiple-indicator factor models

:::

:::{.references}

Mader, N., Arslan, R. C., Schmukle, S. C., & Rohrer, J. M. (2023). Emotional (in)stability: Neuroticism is associated with increased variability in negative emotion after all. *Proceedings of the National Academy of Sciences of the United States of America*, *120(23)*, e2212154120. <https://doi.org/10.1073/pnas.2212154120>

Koslowski, K., & Holtmann, J. (2026). Beneath the Floor: Censored DSEM Models for Analyzing ESM Data with Floor Effects. *Preprint*. <https://osf.io/preprints/psyarxiv/3adgx_v1/>


:::

---

#### Implementing Censoring Models 

::: {.columns}
::: {.column width="47%"}

- Any model can be extended to accommodate censored variables   
- For example, a censored AR(1):

```r
mlts_model(q = 1, censor_left = 1)
```

- where `censor_left = 1` indicates all observations $\leq$ 1 will be assumed as censored
- These observations will be replaced with range-restricted latent parameters during model estimation 
  - Think of it as missing value imputation with the restriction that imputed values can not exceed the censoring threshold
  
:::

::: {.column width="6%"}

:::

::: {.column width="47%"}

:::{.vspace}
:::

:::{.vspace}
:::

#### Caveats 

- Censoring threshold can only be specified globally
  - ergo ALL variables will be affected by censoring
- In multivariate models, when this is not intended, we can use a workaround
  - Assume two variables measured on a scale of 1 to 7
  - To only treat one variable as censored, first shift the scale of that variable
  - Then use the new lower scale bound as the censoring threshold 

```r
Y1 = Y1 - 0.1   # shifts the lower scale end to 0.9

mlts_model(q = 2, censor_left = 0.9) 
```

:::
:::


