
# Live Session II (45 min)

* Session aims:
  * Become familiar with some of the advanced modeling options 
  * Encounter some common issues and implement a fix

* What to do?
  * Work with your own data, OR
  * Try working on the examples provided on the Github-page
    * `live_session/live_session2.R` 
    * We later discuss the solutions
    * Important: For practical reasons, we only use the `bolzen` data frame for all analyses (without `tinterval`). Please do the same but keep in mind that the results may be taken with a large grain of salt.
    * Set `seed = 1234` in `mlts_fit()`, to replicate the exact results
      
--- 

## 1. Bivariate VAR(1) model

::: {.panel-tabset}

### Task

**Model:** 

- Bivariate VAR(1) model with random means

**Variables:** 

- Use repetitive negative thinking (`rnt`) and positive affect composite score (`posaff`) as time series variables

**Data:**

- Use only data of the intervention phase (`phase = "i"` ) and only of subject in the detached mindfulness condition (`group = "dm"`) 

**To explore:**


### Solution

```r
# model specification
model = mlts_model(
  q = 2,
  fix_dynamics = TRUE,
  fix_inno_vars = TRUE
)

# model estimation 
fit_1.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(group == "dm" & phase == "i"), 
  id = "id",
  ts = c("rnt", "posaff"),
  iter = 2000,
  seed = 1234
)
```

### Results 

```{r}
load("live_session/live_session2_fits/Model_1.1_VAR1.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_1.1, digits = 2))
cat(summary_lines[1:31], sep = "\n")  
```



:::

## 2. Between-Level Covariates

::: {.panel-tabset}

### Task

**Model:** 

- Two-level AR(1) model with random means, random AR effects, and random innovation variances
- Add a covariate as predictor for all random effects in the model

**Variables:** 

- Use negative affect composite score (`negaff`) as time series variable
- Use the treatment condition as binary predictor variable 
  - `group` is a character variable that needs to be transformed into a binary-coded (0 and 1) predictor first  

**Data:**

- Use only data of the intervention phase

**Additional:**

- Try requesting standardized regression parameter estimates for the effect of the treatment condition

### Solution

```r
#  Person-level covariate: binary-coded intervention condition 
#  (0 = mindfullness, 1 = detached mindfullness)
bolzen$group_bi = ifelse(bolzen$group == "dm", 1, 0)

# model specification
model = mlts_model(
  q = 1,                   
  ranef_pred = "group_bi" # name of covariate in the data
)

# model estimation 
fit_2.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase == "i"), # subset of the data 
  id = "id",
  ts = c("negaff"),
  iter = 2000,
  seed = 1234,
  # IMPORTANT:
  # For binary covariates, to avoid cluster-mean centering, set
  center_covs = FALSE
)
```

### Results | Model Convergence

```{r}
load("live_session/live_session2_fits/Model_2.1_AR1_with_covariate.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_2.1, flag_signif = TRUE, digits = 2))
cat(summary_lines[1:8], sep = "\n")  # e.g. first 10 
cat("[...]")
```

### Results | Estimates

```{r}
cat(summary_lines[10:33], sep = "\n")  # e.g. first 10 
```

### Results | Std. Estimates
```{r}
#| echo: true
mlts_standardized(fit_2.1)
```

:::


## 3. Prediction of Time-Stable Outcomes

::: {.panel-tabset}

### Task

**Model:** 

- Two-level AR(1) model with random means, random AR effects, and random innovation variances
- Use assessment of trait repetitive negative thinking after the intervention phase (`post_rtq_sum`) as outcome regressed on all random effects in the model AND include baseline assessment of trait repetitive negative (`pre_rqt_sum`) as additional predictor

**Variables:** 

- Use positive affect composite score (`posaff`) as time series variable
- Outcome: `post_rtq_sum`
- Additional predictor: `pre_rqt_sum`

**Data:**

- Use only data of the intervention phase


### Solution

```r
# model specification
model = mlts_model(
  q = 1,                   
  out_pred = "post_rtq_sum",
  out_pred_add_btw = "pre_rtq_sum"
)

# model estimation 
fit_3.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase == "i"), # subset of the data 
  id = "id",
  ts = c("posaff"),
  iter = 2000,
  seed = 1234
)
```

### Results | Model Convergence

```{r}
load("live_session/live_session2_fits/Model_3.1_AR1_with_outcome.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_3.1, flag_signif = TRUE, digits = 2))
cat(summary_lines[1:8], sep = "\n")  # e.g. first 10 
cat("[...]")
```

### Results | Estimates

```{r}
options(width = 200)  # increase from default ~80
cat(summary_lines[10:36], sep = "\n")  # e.g. first 10 
```

### Results | Std. Estimates
```{r}
#| echo: true
mlts_standardized(fit_3.1)
```

:::


## 4. Multiple-Group Models


::: {.panel-tabset}

### Task

**Model:** 

- Multiple-group two-level AR(1) model with random means, random AR effects, and random innovation variances

**Variables:** 

- Use negative affect composite score (`negaff`) as time series and treatment condition (`group`) as grouping variable

**Data:**

- Use only data of the intervention phase

**To explore:**

- Run comparisons to test group-differences in average parameters
  - Access the posterior draws of parameters of the fitted model (`fit$posteriors`)
  - Compute parameter difference scores for each MCMC draw (e.g., average AR in group 1 - average AR in group 2)
  - Calculate summary statistics of differences scores for statistical inference 
    - mean or median as point estimates for average group differences, and 95% credible intervals (using `quantile()`)
  - Are the results consistent with "2. Between-Level Covariates"?

### Solution

```r
# model specification
model = mlts_model(
  q = 1,
  group = 2
)

# model estimation 
fit_4.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase == "i"), # subset of the data 
  id = "id",
  ts = c("negaff"),
  group = "group",  # new argument: pass name of grouping variable
  iter = 2000,
  seed = 1234
)
```

### Results | Model Convergence

```{r}
load("live_session/live_session2_fits/Model_4.1_multiple_group.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_4.1, digits = 2))
cat(summary_lines[1:11], sep = "\n")  # e.g. first 10 
cat("[...]")
```

### Results | Estimates

::: {.column width="50%"}

```{r}
cat(do.call(rbind,lapply(summary_lines[13], function(x){str_sub(x, start = 1, end = 55)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[14:21], function(x){str_sub(x, start = 1, end = 35)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[22:27], function(x){str_sub(x, start = 1, end = 40)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[28:31], function(x){str_sub(x, start = 1, end = 45)})), sep = "\n")
```

:::

::: {.column width="50%"}

```{r}
cat(do.call(rbind,lapply(summary_lines[33], function(x){str_sub(x, start = 1, end = 55)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[34:41], function(x){str_sub(x, start = 1, end = 35)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[42:47], function(x){str_sub(x, start = 1, end = 40)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[48:51], function(x){str_sub(x, start = 1, end = 45)})), sep = "\n")
```

:::

### Comparisons I 

**Difference in negative affect inertia**

First, calculate the differences in group-specific estimates of the average AR 
```{r}
#| echo: true
# Compute difference scores based on post-warmup MCMC samples:
diff = as.vector(fit_4.1$posteriors[,,"dm: phi(1)_11"] - 
                 fit_4.1$posteriors[,,"gi: phi(1)_11"])
```

Then, calculate summary statistics across this posterior distribution of this difference score
```{r}
#| echo: true
# Get summary statistics across MCMC draws
round(c("M" = mean(diff), "SD" = sd(diff), 
        quantile(diff, probs = c(.025, .975))),3)
```

**Compare to results of Model 2.1**
As a small validity check, we can compare the above results with the results of Model 2.1
where `group_bi` was used as a predictor of differences in AR effects.
To this end, we inspect the results of the respective between-level regression weight:
```{r}
#| echo: true
par = "b_phi(1)_11.ON.group_bi" # read: phi regressed on group_bi
fit_2.1$pop.pars.summary[fit_2.1$pop.pars.summary$Param == par, c("mean", "sd", "2.5%", "97.5%")]
```


### Comparisons II

Multiple group models allow comparisons beyond differences in average (fixed) effects:  

- Differences in random effect variances (SDs) and random effect correlations

Same procedure: Obtain difference scores first, then compute summary statistics. 
  - For example, correlation between means and log innovation variances

```{r}
#| echo: true
diff = as.vector(fit_4.1$posteriors[,,"dm: r_mu_1.ln.sigma2_1"] - 
                 fit_4.1$posteriors[,,"gi: r_mu_1.ln.sigma2_1"])

# Get summary statistics across MCMC draws
round(c("M" = mean(diff), "SD" = sd(diff), 
        quantile(diff, probs = c(.025, .975))),3)
```


### Results Plot 
`mlts_plot` now plots the group-specific estimates

```{r}
#| echo: true
#| out-width: '60%'
mlts_plot(fit_4.1)
```

:::


## 5. Interaction Effects on the Dynamic Within-Level

::: {.panel-tabset}

### Task

**RQ1**: Does average levels of repetitive negative thinking (`rnt`) change between baseline and intervention (`phase`) period?

**RQ2**: Does the average inertia of repetitive negative thinking (`rnt`) change between baseline and intervention (`phase`) period?

**Model:** 

- Random-intercept moderated VAR(1) model
- Time-varying covariate should be treated as exogenous variable

**Variables:** 

- `rnt` and `phase_bi` (needs to be created) 

**Data:**

- Use the complete data

**To Explore:**

- How can we test whether the amount of change in the dynamics of RNT between study and intervention phase depend on the treatment condition?   

### Solution

::: {.column width="48%"}
```r
bolzen$phase_bi = ifelse(bolzen$phase == "b", 0,1)

# model specification
model <- mlts_model(
  q = 2,                          
  is_exogenous = 1,               
  incl_t0_effects = "phi(0)_21",  
  fixef_zero = "phi(1)_21",       
  inno_covs_zero = TRUE,            
  
  # include interaction effect:
  incl_interaction_effects = "phi(i)_2.2(1)1(0)",
  
  # assume constant regression effects and innovation variances
  fix_dynamics = T,
  fix_inno_vars = T
)
```
:::

::: {.column width="4%"}

:::

::: {.column width="48%"}
```r
# model estimation 
fit_5.1 = mlts_fit(
  model = model,
  data = bolzen, 
  id = "id",
  ts = c("phase_bi", "rnt"),
  iter = 2000,
  seed = 1234,
)
```
:::

### Results 

```{r}
load("live_session/live_session2_fits/Model_5.1_interaction.Rdata")
```


::: {.column width="65%"}

```{r}
summary_lines <- capture.output(summary(fit_5.1, flag_signif = TRUE, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 

```

```{r}
cat(summary_lines[11:22], sep = "\n")  # e.g. first 10 
```
:::

::: {.column width="35%"}
**Parameter interpretation**

- mu_2: Average levels of RNT in the baseline phase (i.e., `phase_bi = 0`)
- phi(0)_21: Change in average levels of RNT from baseline to intervention phase (i.e., `phase_bi = 1`)
- phi(1)_22: Average AR effect in the baseline phase (i.e., `phase_bi = 0`)
- phi(i)_2.2(1)1(0): Change in average AR effect from baseline to intervention phase (i.e., `phase_bi = 1`)

:::

### To Explore

Using the same restrictions as before (no random effects for AR or interaction effects), we can not directly include `group` as a between-level predictor. But, we can fit a multiple-group model: 

::: {.column width="48%"}
```r
# model specification
model <- mlts_model(
  q = 2,                          
  is_exogenous = 1,               
  incl_t0_effects = "phi(0)_21",  
  fixef_zero = "phi(1)_21",       
  inno_covs_zero = TRUE,            
  incl_interaction_effects = "phi(i)_2.2(1)1(0)",
  fix_dynamics = T,
  fix_inno_vars = T,
  group = 2                 # !
)
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

```r
# model estimation 
fit_5.2 = mlts_fit(
  model = model,
  data = bolzen, 
  id = "id",
  ts = c("phase_bi", "rnt"),
  group = "group",         # ! 
  iter = 2000,
  seed = 1234,
)
```
:::

### To Explore | Results

```{r}
load("live_session/live_session2_fits/Model_5.2_interaction_by_group.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_5.2, flag_signif = TRUE, digits = 2))
```

```{r}
cat(summary_lines[13:40], sep = "\n")  # e.g. first 10 
```


:::


## 6. Multiple-Indicator Factor Models

::: {.panel-tabset}

### Task

**Model:** 

- Three-indicator AR(1) model with constant AR effect and innovation variance
- Assume random indicator-specific person means (no common between-level factor)

**Variables:** 

- Use items *anxious* (`aengstlich`), *nervous* (`nervoes`), and *relaxed (recoded)* (`entspannt`) as indicators of a common latent construct (tense arousal) 

**Data:**

- For faster estimation, use only data of the detached mindfulness group (`group = "dm"`)


### Solution

```r
# first, rescale relaxed item
bolzen$rlx_r = 8 - bolzen$entspannt

# model specification
model = mlts_model(
  q = 1,   # one common latent construct
  p = 3,   # measured by three manifest indicators
  fix_dynamics = TRUE,
  fix_inno_vars = TRUE,
  btw_factor = FALSE
)

# run again without this subject 62: 
fit_6.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(group == "dm"),
  id = "id",
  ts = c("aengstlich", "nervoes", "rlx_r"),
  iter = 3000,
  seed = 1234
)
```

### Results | Model Convergence

```{r}
load("live_session/live_session2_fits/Model_6.1_ThreeIndicator_AR1.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_6.1, digits = 2))
cat(summary_lines[1:8], sep = "\n")  # e.g. first 10 
cat("[...]")
```

### Results | Estimates

```{r}
cat(summary_lines[10:38], sep = "\n")  # e.g. first 10 
```


:::


## 7. Censoring Models

::: {.panel-tabset}

### Task

**Model:** 

- Censored two-level AR(1) model with random means, random AR effects, and random innovation variances 

**Variables:** 

- Use negative affect composite score (`negaff`) as time series variable

**Data:**

- Use the complete dataset

**To explore:**

- Calculate the amount of floor effects
- Fit standard AR(1) model and compare results 

### Solution | Censored AR(1)

```r
describe_ts(data = bolzen, ts = "negaff")
# > 10.9% of non-missing observations at the lower scale bound!

# model specification
model = mlts_model(
  q = 1,
  censor_left = LB,  # equals 1 (see above)
)

# model estimation 
fit_7.1 = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = "negaff",
  iter = 2000
)
```

### Solution | AR(1)

```r
# model specification
model = mlts_model(
  q = 1
)

# model estimation 
fit_7.2 = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = "negaff",
  iter = 2000
)
```

### Result Comparison 

```{r}
load("live_session/live_session2_fits/Model_7.1_CensoredAR1.Rdata")
load("live_session/live_session2_fits/Model_7.2_AR1.Rdata")

summary_lines <- capture.output(summary(fit_7.1, flag_signif = TRUE, digits = 2))
summary_lines2 <- capture.output(summary(fit_7.2, flag_signif = TRUE, digits = 2))
```


::: {.column width="50%"}

#### Censored

```{r}
cat(do.call(rbind,lapply(summary_lines[11:16], function(x){str_sub(x, start = 1, end = 35)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[17:22], function(x){str_sub(x, start = 1, end = 40)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines[23:27], function(x){str_sub(x, start = 1, end = 45)})), sep = "\n")
```

:::

::: {.column width="50%"}

#### Standard

```{r}
cat(do.call(rbind,lapply(summary_lines2[11:16], function(x){str_sub(x, start = 1, end = 35)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines2[17:22], function(x){str_sub(x, start = 1, end = 40)})), sep = "\n")
cat(do.call(rbind,lapply(summary_lines2[23:27], function(x){str_sub(x, start = 1, end = 45)})), sep = "\n")
```

:::

### Model Comparison | PPCs
Using functions implemented in the `bayesplot`-package

::: {.column width="50%"}
1. Get posterior predictive samples according to both models
```{r, echo = TRUE, eval = F}
y_reps_cens = mlts_posterior_sample(
  fit = fit_7.1, n_draws = 20, draw_person_pars = T)
y_reps_ar   = mlts_posterior_sample(
  fit = fit_7.2, n_draws = 20, draw_person_pars = T)
```

```{r}
#save(y_reps_cens, file = "../live_session/live_session2_fits/Model_7.1_CensoredAR1_yreps.Rdata")
#save(y_reps_ar,   file = "../live_session/live_session2_fits/Model_7.2_AR1_yreps.Rdata")

load(file = "live_session/live_session2_fits/Model_7.1_CensoredAR1_yreps.Rdata")
load(file = "live_session/live_session2_fits/Model_7.2_AR1_yreps.Rdata")
```

2. Store the observations as a vector
```{r, echo = TRUE}
y = fit_7.1$data$negaff
```

3. If missing values are present: 
```{r, echo = TRUE}
# index and remove missing values from y and y_rep
take = !is.na(y)     
y = y[take]
y_reps_cens <- y_reps_cens$negaff[,take]
y_reps_ar   <- y_reps_ar$negaff[,take]
```
:::

::: {.column width="50%"}

4. Run graphical checks using the replicated datasets 
```{r, echo = TRUE} 
#| layout-ncol: 2
#| out-height: "55%"

# AR(1) model
bayesplot::ppc_dens_overlay(y = y, yrep = y_reps_ar) + theme_classic() +
  labs(title = "AR(1)") +
  scale_x_continuous(limits = c(1, 7))
# censored AR(1) model
bayesplot::ppc_dens_overlay(y = y, yrep = y_reps_cens) + theme_classic() +
  labs(title = "Censored AR(1)") +
  scale_x_continuous(limits = c(1, 7))
```
:::

:::


