
# Model Extensions I

```{r}
dir.create("sections/06-model_extensions/")

# load all fitted models
for(i in Sys.glob("sections/06-model_extensions/*.Rdata")){
  load(i)
}
```

---

### Vector Autoregressive Model

- Allows estimation of time-lagged effects between multiple variables
  - Autoregressive and cross-lagged (CL) effects
- In multivariate models, the number of random effects can quickly increase
  - Estimation times can quickly increase, especially for large *N*
  - Yielding acceptable model convergence criteria can be much challenging
  
- We will illustrate this for the positive and negative affect composite scores

---

:::: {.columns}

::: {.column width="35%"}

#### Decomposition

Applied to each variable:
$$
\begin{matrix}
Y_{1it}^{w} &= Y_{1it} - \mu_{1i} \\
Y_{2it}^{w} &= Y_{2it} - \mu_{2i} \\
\end{matrix}
$$ 

#### Within-level
$$
\begin{matrix}
Y_{1it}^w &= \phi_{11i} Y_{1i(t-1)}^{w} + \phi_{12i} Y_{2i(t-1)}^{w} + \zeta_{1it} \\
Y_{2it}^w &= \phi_{21i} Y_{1i(t-1)}^{w} + \phi_{22i} Y_{2i(t-1)}^{w} + \zeta_{2it} \\
\end{matrix}
$$

$$
\begin{matrix}
\text{with}\ \zeta_{1it} \sim N(0, \sigma_{\zeta 1i}^2) \\
\text{with}\ \zeta_{2it} \sim N(0, \sigma_{\zeta 2i}^2)
\end{matrix}
$$ 

#### Between-level
$$
\begin{matrix}
\mu_{1i}  = \gamma_{\mu1}  + \upsilon_{\mu1i} \\
\mu_{2i}  = \gamma_{\mu1}  + \upsilon_{\mu1i} \\
\phi_{11i} = \gamma_{\phi11} + \upsilon_{\phi 11i} \\
\phi_{12i} = \gamma_{\phi12} + \upsilon_{\phi 12i} \\
[...] \\
ln(\sigma_{\zeta 2i}^2) = \gamma_{ln(\sigma^2_{\zeta2})} + \upsilon_{ln(\sigma^2_{\zeta 2i})}
\end{matrix}
$$ 

$$
\text{with}\ \mathbf{\upsilon} \sim MVN(\mathbf{0},\mathbf{\Sigma})
$$ 

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

#### Model Specification

```{r, echo = T}
model = mlts_model(
  q = 2,                 # number of time-varying constructs
  inno_covs_zero = TRUE  # assume independent innovations (r = 0)
  )
```


```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.8, scale_within_inno = 0.22,
           cex_title = 0.8, scale_between = 0.32, asp_w_b = 0.4)
```

:::
::::

--- 

### Fixed Effect Parameters in the `model`-object
```{r}
#| echo: true
model[1:8,c("Type","Param","Param_Label")]
```

- `phi(1)_11` and `phi(1)_22` are the AR effects of variable 1 and 2
- `phi(1)_21` and `phi(1)_12` are the first-order CL effect between variables 1 and 2. The first number indicates the DV (outcome) and the second number the IV (predictor) 

- The order of the variables (1 or 2) will be determined by the order in which the variable names are passed to the `ts`-argument in `mlts_fit()`. 

--- 

## Model Estimation | A Detailled Workflow

::: {.column width="55%"}

- For ESM datasets of moderate sample size (N ~ 100 & T ~ 70), estimating VAR(1) models with random effects for all time-lagged effects and innovation variances can be difficult
  - Insufficient model convergence indices & long estimation times 

- Usually the result of model misfit 
  - Known issue: Parameter estimates close to boundary values
    - Low random effect variances
    - Random effect correlations close to 1 (or -1)
  - Preliminary estimation runs (with ~ 500-1000 iterations) can help to diagnose model misfit at an early stage 

:::

. . .

::: {.column width="5%"}

:::

::: {.column width="40%"}
```{r}
#| eval: false
#| echo: true
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 500)
```
:::

---

### Preliminary Run | Results

:::: {.columns}

::: {.column width="45%"}

```{r, eval = F}
fit_VAR = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR, file = "06-model_extensions/1.1_VAR_500.Rdata")

fit_VAR_l = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  iter = 2000,
  seed = 1234)

save(fit_VAR_l, file = "06-model_extensions/1.1_VAR_2000.Rdata")

```

```{r}
load("sections/06-model_extensions/1.1_VAR_500.Rdata")
load("sections/06-model_extensions/1.1_VAR_2000.Rdata")
```

```{r}
#| eval: false
#| echo: true
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 500)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time: Longest chain took 14.9 minutes


:::
::::

---

### Preliminary Run | Results for `iter = 2000`

:::: {.columns}

::: {.column width="45%"}

```{r}
#| eval: false
#| echo: true
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR_l = mlts_fit(
  model = model,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR_l, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time: Longest chain took 36.5 minutes

:::
::::

--- 

### Running Model With Constraints

```{r, eval = F}
model_r = mlts_model(q = 2, inno_covs_zero = T, 
                     ranef_zero = "phi(1)_21")

fit_VAR_r = mlts_fit(
  model = model_r,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_r, file = "06-model_extensions/1.1_VAR_r1_500.Rdata")

fit_VAR_r_l = mlts_fit(
  model = model_r,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_r_l, file = "06-model_extensions/1.1_VAR_r1_2000.Rdata")

model_r2 = mlts_model(q = 2, inno_covs_zero = T, 
                     ranef_zero = c("phi(1)_21", "phi(1)_12"))

fit_VAR_r2 = mlts_fit(
  model = model_r2,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_r2, file = "06-model_extensions/1.1_VAR_r2_500.Rdata")

fit_VAR_r2_l = mlts_fit(
  model = model_r2,
  data = bolzen,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_r2_l, file = "06-model_extensions/1.1_VAR_r2_2000.Rdata")
```


::: {.column width="25%"}

:::

::: {.column width="7%"}
**iter**
:::

::: {.column width="7%"}
**Time**
:::

::: {.column width="20%"}
**Diagnostics**
:::

::: {.column width="41%"}
**Crit. Parameters**
:::


```r
mlts_model(q = 2, inno_covs_zero = T)
```

::: {.column width="25%"}

:::

::: {.column width="7%"}
500
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun <- function(fit){
  max.rhat = round(max(fit$pop.pars.summary$Rhat),3)
  min.ess_bulk = min(fit$pop.pars.summary$Bulk_ESS)
  min.ess_tail = min(fit$pop.pars.summary$Tail_ESS)
  n_div  = rstan::get_num_divergent(fit$stanfit)
  cat(paste0(
    "Max Rhat     = ", max.rhat, "\n",
 #   "Min Tail ESS = ", min.ess_tail, "\n",
#    "Min Bulk ESS = ", min.ess_bulk, "\n",
    "N divergent  = ", n_div
  ))
}
diag_fun(fit_VAR)
```
:::

::: {.column width="41%"}
```{r}
fun_crit <- function(fit){
  cols = c("Param", "Rhat","Bulk_ESS", "Tail_ESS")
  a = which(fit$pop.pars.summary$Tail_ESS == min(fit$pop.pars.summary$Tail_ESS))
  b = which(fit$pop.pars.summary$Bulk_ESS == min(fit$pop.pars.summary$Bulk_ESS))
  c = which(fit$pop.pars.summary$Rhat == max(fit$pop.pars.summary$Rhat))
  fit$pop.pars.summary[unique(c(a,b,c)), cols]  
}

fun_crit(fit_VAR)
```
:::

<!-- new row -->

::: {.column width="25%"}

:::

::: {.column width="7%"}
2000
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_l$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_l)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_l)
```
:::

<!-- new row -->

:::{.vspace}
:::

```r
mlts_model(q = 2, inno_covs_zero = T, ranef_zero = "phi(1)_21")
```


::: {.column width="25%"}

:::

::: {.column width="7%"}
500
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r)
```
:::

<!-- new row -->

::: {.column width="25%"}

:::

::: {.column width="7%"}
2000
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r_l$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r_l)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r_l)
```
:::

<!-- new row -->


:::{.vspace}
:::

```r
mlts_model(q = 2, inno_covs_zero = T, ranef_zero = c("phi(1)_21","phi(1)_12"))
```


::: {.column width="25%"}

:::

::: {.column width="7%"}
500
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r2$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r2)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r2)
```
:::

<!-- new row -->

::: {.column width="25%"}

:::

::: {.column width="7%"}
2000
:::

::: {.column width="7%"}
```{r}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_r2_l$stanfit),1,sum))/60,2))
```
:::

::: {.column width="20%"}
```{r}
diag_fun(fit_VAR_r2_l)
```
:::

::: {.column width="41%"}
```{r}
fun_crit(fit_VAR_r2_l)
```
:::

---

### Take-Away

- Estimation times can quickly increase, especially when the model does not fit the data
  - here: random effect variances of the CL effects 
- Preliminary test runs with less iterations can usually detect misfit early on 
  - Strong recommended especially for complex models 
- When the model fits the data, running more iterations is usually less costly 


---

### Take-Away

- Estimation times can quickly increase, especially when the model does not fit the data
  - here: random effect variances of the CL effects 
- Preliminary test runs with less iterations can usually detect misfit early on 
  - Strong recommended especially for complex models 
- When the model fits the data, running more iterations is usually less costly 


:::{.vspace}
:::

:::{.large}
**Are we Done here?**
:::


---

### Are we Done here?

- Let's recap, why did the unrestricted VAR(1) model fail?
- Any suggestions what might change the results?

```{r}
summary_lines <- capture.output(summary(fit_VAR_r2_l, digits = 2))
cat("[...]")
cat(summary_lines[11:21], sep = "\n")  # e.g. first 10 
cat("[...]")
```

---

### Refitting the Model with `tinterval`

```{r, eval = F}
fit_VAR_tint = mlts_fit(
  model = model,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_tint, file = "06-model_extensions/1.1_VAR_tint_500.Rdata")

fit_VAR_tint_l = mlts_fit(
  model = model,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_tint_l, file = "06-model_extensions/1.1_VAR_tint_2000.Rdata")


fit_VAR_tint_r = mlts_fit(
  model = model_r,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234)

save(fit_VAR_tint_r, file = "06-model_extensions/1.1_VAR_tint_r_500.Rdata")

fit_VAR_tint_r_l = mlts_fit(
  model = model_r,
  data = bolzen_grid,
  time = "int_time_spaced",
  tinterval = 1,
  max_NA_seq = 4,
  id = "id",
  ts = c("posaff", "negaff"),
  seed = 1234,
  iter = 2000)

save(fit_VAR_tint_r_l, file = "06-model_extensions/1.1_VAR_tint_r_2000.Rdata")

```

:::: {.columns}

::: {.column width="45%"}

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: 8-10
model = mlts_model(q = 2, inno_covs_zero = T)

fit_VAR_tint_l = mlts_fit(
  model = model,
  data = bolzen_grid,
  id = "id",
  ts = c("posaff", "negaff"),
  tinterval = 1,
  time = "int_time_spaced",
  max_NA_seq = 4,
  seed = 1234,
  iter = 2000)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR_tint_l, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time:
```{r, eval = F}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_tint_l$stanfit),1,sum))/60,2))
```
Longest chain took 16.1 minutes


:::
::::

---

### Refitting the Model with `tinterval` II

:::: {.columns}

::: {.column width="45%"}

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: 1-13|2
model_r = mlts_model(q = 2, inno_covs_zero = T, 
                     ranef_zero = "phi(1)_21")

fit_VAR_tint_r_l = mlts_fit(
  model = model_r,
  data = bolzen_grid,
  id = "id",
  ts = c("posaff", "negaff"),
  tinterval = 1,
  time = "int_time_spaced",
  max_NA_seq = 4,
  seed = 1234,
  iter = 2000)
```

```{r}
summary_lines <- capture.output(summary(fit_VAR_tint_r_l, digits = 2))
cat(summary_lines[1:9], sep = "\n")  # e.g. first 10 
cat("[...]")

```

:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

```{r}
cat(summary_lines[11:32], sep = "\n")  # e.g. first 10 
```

:::{.vspace}
:::

Check estimation time:
```{r, eval = F}
cat(round(max(chain_sec = apply(rstan::get_elapsed_time(fit_VAR_tint_r_l$stanfit),1,sum))/60,2))
```
Longest chain took 15.1 minutes


:::
::::



---



### Innovation Covariance - Three modeling options

:::: {.columns}

::: {.column width="30%"}

1. Independent: 
```{r}
#| code-line-numbers: 3-3
#| eval: false
#| echo: true
mlts_model(
  q = 2,
  inno_covs_zero = TRUE
  )
```


:::

::: {.column width="5%"}
:::

::: {.column width="30%"}

2. Constant innovation correlation 
```{r}
#| code-line-numbers: 3-4
#| eval: false
#| echo: true
mlts_model(
  q = 2,
  inno_covs_zero = FALSE,  # default
  fix_inno_covs = TRUE     # default
  )
```

:::

::: {.column width="5%"}

:::

::: {.column width="30%"}

3. Random Innovation covariance 
```{r}
#| code-line-numbers: 3-5
#| eval: false
#| echo: true
mlts_model(
  q = 2,
  inno_covs_zero = FALSE,
  fix_inno_covs = FALSE,
  inno_covs_dir = "pos" # or "neg"
  )
```

:::
::::

:::: {.columns}

::: {.column width="30%"}
Implies:

$$
\begin{bmatrix}
\zeta_{1it} \\
\zeta_{2it} 
\end{bmatrix} 
\sim MVN 
\left(
\begin{bmatrix}
0 \\ 0
\end{bmatrix},
\begin{bmatrix}
\sigma_{\zeta 1i}^2 & 0 \\ 
0 & \sigma_{\zeta 2i}^2
\end{bmatrix}
\right)
$$ 
:::

::: {.column width="5%"}

:::

::: {.column width="30%"}
Implies:

$$
\begin{bmatrix}
\zeta_{1it} \\
\zeta_{2it} 
\end{bmatrix} 
\sim MVN 
\left(
\begin{bmatrix}
0 \\ 0
\end{bmatrix},
\begin{bmatrix}
\sigma_{\zeta 1i}^2 & 0 \\ 
\sigma_{\zeta 12i}^2 & \sigma_{\zeta 2i}^2
\end{bmatrix}
\right)
$$ 
where
$$
\sigma_{\zeta 12i}^2 = \rho_{\zeta12} * \sigma_{\zeta 1i} * \sigma_{\zeta 2i}
$$
$\rho_{\zeta12}$ without index $i$ --> constant across clusters

:::

::: {.column width="5%"}

:::

::: {.column width="30%"}
Implies:
$$
\begin{matrix}
\zeta_{1it} = 1 * \eta_{\zeta,it} + \delta_{1it} \\
\zeta_{2it} = 1 * \eta_{\zeta,it} + \delta_{2it} \\
\end{matrix} 
$$

$$
\begin{matrix}
\delta_{1it} \sim N(0, \sigma_{\delta 1i}^2)\\
\delta_{2it} \sim N(0, \sigma_{\delta 2i}^2)\\
\eta_{\zeta,it} \sim N(0, \sigma_{\eta \zeta i}^2)
\end{matrix} 
$$ 

:::
::::

---


### Between-level Covariates  

Person-level variables as predictor of individual differences in any of the dynamic model parameters:  

:::: {.columns}

::: {.column width="35%"}

#### Decomposition

- Latent Person-mean-centering:

$$
Y_{it}^{w} = Y_{it} - \mu_{i}
$$ 

#### Within-level
$$
Y_{it}^w = \phi_i Y_{i(t-1)}^{w} + \zeta_{it} \\
\text{with}\ \zeta_{it} \sim N(0, \sigma_{\zeta i}^2)
$$ 


#### Between-level
$$
\begin{matrix}
\mu_i  &=& \gamma_{\mu}  &+& b_{\mu}Group_i  &+& \upsilon_{\mu i} \\
\phi_i &=& \gamma_{\phi} &+& b_{\phi}Group_i  &+& \upsilon_{\phi i} \\
ln(\sigma_{\zeta i}^2) &=& \gamma_{ln(\sigma^2_\zeta)} &+& b_{ln(\sigma_\zeta^2)}Group_i  &+& \upsilon_{ln(\sigma^2_{\zeta i})}
\end{matrix}
$$ 

$$
\text{with}\ \mathbf{\upsilon} \sim MVN(\mathbf{0},\mathbf{\Sigma})
$$ 

:::

::: {.column width="10%"}
<!-- Add some space between columns -->
:::

::: {.column width="55%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 1, 
  ranef_pred = "group_bi"  # name of the variable in the dataset
  )
```

We use the dummy-coded treatment condition variable (i.e., `group_bi`).

```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3
#| fig-dpi: 1000
#| fig-align: left


mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.25,
           asp_w_b = 0.6,
           scale_within_inno = 0.3,
           cex_title = 0.8)
```

:::
::::

---

### Important

  - By default between-level covariates will be grand-mean centered to allow a clearer interpretation of the dynamic model parameters 
  - For binary predictor variables, you can turn off this behavior by setting ` mlts_fit(..., center_covs = FALSE)`

---

### Prediction of Time-Stable Outcomes

Person-level variables as outcomes to be predicted by person-specific parameters: 

:::: {.columns}

::: {.column width="55%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 1, 
  out_pred = "Out" # name of the variable in the dataset
  )
```

```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left


mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.2,
           asp_w_b = 0.6,
           scale_within_inno = 0.3,
           cex_title = 0.8)
```
:::

::: {.column width="10%"}
<!-- Add some space between columns -->
:::

::: {.column width="35%"}
**Further options**

- For multiple outcomes: 
  - `out_pred = c("Out1", "Out2")`
- Specific paths: 
  - `list(c("Out" = "mu_1"))`
  - Uses only individual mean levels (`mu_1`) as predictor
- Include observed binary or continuous person-level variables: 
  - `out_pred_add_btw = "Z"`, or:
  - `list(c("Out" = "mu_1", "Z"))`
  - where `"Z"` refers to a variable in the dataset


:::

::::

---

