
# Live Session II Part A (30 min)

* Session aims:
  * Become familiar with some of the advanced modeling options 
  * Encounter some common issues and implement a fix

* What to do?
  * Work with your own data, OR
  * Try working on the examples provided on the Github-page
    * `live_session/live_session2.R` 
    * We later discuss the solutions
    * Important: For practical reasons, we only use the `bolzen` data frame for all analyses (without `tinterval`). Please do the same but keep in mind that the results may be taken with a large grain of salt.
    * Set `seed = 1234` in `mlts_fit()`, to replicate the exact results
      
--- 

## 1. Bivariate VAR(1) model

::: {.panel-tabset}

### Task

**Model:** 

- Bivariate VAR(1) model with random means

**Variables:** 

- Use repetitive negative thinking (`rnt`) and positive affect composite score (`posaff`) as time series variables

**Data:**

- Use only data of the intervention phase (`phase = "i"` ) and only of subject in the detached mindfulness condition (`group = "dm"`) 



### Solution

```r
# model specification
model = mlts_model(
  q = 2,
  fix_dynamics = TRUE,
  fix_inno_vars = TRUE
)

# model estimation 
fit_1.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(group == "dm" & phase == "i"), 
  id = "id",
  ts = c("rnt", "posaff"),
  iter = 2000,
  seed = 1234
)
```

### Results 

```{r}
load("live_session/live_session2_fits/Model_1.1_VAR1.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_1.1, digits = 2))
cat(summary_lines[1:31], sep = "\n")  
```



:::

## 2. Between-Level Covariates

::: {.panel-tabset}

### Task

**Model:** 

- Two-level AR(1) model with random means, random AR effects, and random innovation variances
- Add a covariate as predictor for all random effects in the model

**Variables:** 

- Use negative affect composite score (`negaff`) as time series variable
- Use the treatment condition as binary predictor variable 
  - `group` is a character variable that needs to be transformed into a binary-coded (0 and 1) predictor first  

**Data:**

- Use only data of the intervention phase

**Additional:**

- Try requesting standardized regression parameter estimates for the effect of the treatment condition

### Solution

```r
#  Person-level covariate: binary-coded intervention condition 
#  (0 = mindfullness, 1 = detached mindfullness)
bolzen$group_bi = ifelse(bolzen$group == "dm", 1, 0)

# model specification
model = mlts_model(
  q = 1,                   
  ranef_pred = "group_bi" # name of covariate in the data
)

# model estimation 
fit_2.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase == "i"), # subset of the data 
  id = "id",
  ts = c("negaff"),
  iter = 2000,
  seed = 1234,
  # IMPORTANT:
  # For binary covariates, to avoid cluster-mean centering, set
  center_covs = FALSE
)
```

### Results | Model Convergence

```{r}
load("live_session/live_session2_fits/Model_2.1_AR1_with_covariate.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_2.1, flag_signif = TRUE, digits = 2))
cat(summary_lines[1:8], sep = "\n")  # e.g. first 10 
cat("[...]")
```

### Results | Estimates

```{r}
cat(summary_lines[10:33], sep = "\n")  # e.g. first 10 
```

### Results | Std. Estimates
```{r}
#| echo: true
mlts_standardized(fit_2.1)
```

:::


## 3. Prediction of Time-Stable Outcomes

::: {.panel-tabset}

### Task

**Model:** 

- Two-level AR(1) model with random means, random AR effects, and random innovation variances
- Use assessment of trait repetitive negative thinking after the intervention phase (`post_rtq_sum`) as outcome regressed on all random effects in the model AND include baseline assessment of trait repetitive negative (`pre_rqt_sum`) as additional predictor

**Variables:** 

- Use positive affect composite score (`posaff`) as time series variable
- Outcome: `post_rtq_sum`
- Additional predictor: `pre_rqt_sum`

**Data:**

- Use only data of the intervention phase


### Solution

```r
# model specification
model = mlts_model(
  q = 1,                   
  out_pred = "post_rtq_sum",
  out_pred_add_btw = "pre_rtq_sum"
)

# model estimation 
fit_3.1 = mlts_fit(
  model = model,
  data = bolzen %>% filter(phase == "i"), # subset of the data 
  id = "id",
  ts = c("posaff"),
  iter = 2000,
  seed = 1234
)
```

### Results | Model Convergence

```{r}
load("live_session/live_session2_fits/Model_3.1_AR1_with_outcome.Rdata")
```

```{r}
summary_lines <- capture.output(summary(fit_3.1, flag_signif = TRUE, digits = 2))
cat(summary_lines[1:8], sep = "\n")  # e.g. first 10 
cat("[...]")
```

### Results | Estimates

```{r}
options(width = 200)  # increase from default ~80
cat(summary_lines[10:36], sep = "\n")  # e.g. first 10 
```

### Results | Std. Estimates
```{r}
#| echo: true
mlts_standardized(fit_3.1)
```

:::


