
# Model Extensions II

```{r}
dir.create("sections/06-model_extensions/")

# load all fitted models
for(i in Sys.glob("sections/06-model_extensions/*.Rdata")){
  load(i)
}
```

---

### Multiple-Group Models

:::{.large}

- Simultaneous estimation of the *same dynamic model* *separately for a specified number of groups* 

- As shown before, differences in model parameters can be estimated by adding the group-variable as predictor of individual-specific effects. 
  - **But**: only possible for random effects.

- Multiple-group models: Allow estimation of group-specific model parameters 
  - Differences in average (i.e. fixed) effects, random effect variance, random effect correlation, 
  innovation correlations

- Possible scenarios where this is relevant:
  - Relatively low number of measurements which would require estimating some model parameters as constant but group-specific differences therein are of interest

:::  

---

### Multiple-Group Models

::: {.column width="50%"}
#### Model Specification 
```{r}
#| echo: true
VAR1_g <- mlts_model(
  q = 2,
  group = 2,  
  fix_dynamics = TRUE,
  fix_inno_vars = TRUE,
  )
```

#### Model Estimation
```{r}
#| echo: true
#| eval: false
fit <- mlts_fit(
  model = VAR1_g,
  data = bolzen %>% filter(phase == "i"),
  id = "id",
  ts = c("negaff","posaff"),
  group = "group",
  iter = 2000
  )
```
```{r}
#save(fit, file = "06-model_extensions/4.1_MG.Rdata")
load("sections/06-model_extensions/4.1_MG.Rdata")
```
:::

::: {.column width="50%"}
#### Results
```{r}
summary_lines <- capture.output(summary(fit, digits = 2))
cat(summary_lines[1:13], sep = "\n")  
```
:::

---

### Multiple-Group Models

#### Results
::: {.column width="50%"}

```{r}
cat(summary_lines[14:35], sep = "\n")  
```

:::

::: {.column width="50%"}
```{r}
cat(summary_lines[37:58], sep = "\n")  
```
:::


---

### Multiple-Group Models | Group-Comparisons

- Based on the group-specific estimates, we can run comparison tests
- The procedure to conduct these tests is the same for any model parameter (also random effect variances or random effect correlations)
- For example, are there group-differences in the average CL effect of `poasaff` at $t-1$ on `negaff` at $t$ (i.e., `phi(1)_12`).

- We first extract the posterior draws of the group-specific estimates:
```{r}
#| echo: true
CL_dm = fit$posteriors[,,"dm: phi(1)_12"]
CL_gi = fit$posteriors[,,"gi: phi(1)_12"]
```

- Then, calculate the differences between the group-specific estimates for each draw:
```{r}
#| echo: true
Diff = as.vector( CL_dm - CL_gi )
```

- Now, we can calculate summary statistics: 
  - such as the difference parameters' posterior mean, 
```{r}
#| echo: true
mean(Diff)
```
  - or a 95% CI to test potential differences with $\alpha = .05$:
```{r}
#| echo: true
quantile(Diff, probs = c(0.025, 0.975))
```

---

### Interaction Effects on the Dynamic Within-Level


- Thus far, we assumed *AR*- or *CL*-effects to be invariant over time

- This assumption can be relaxed or tested, e.g., using **time-varying effects models** [@Bringmann2017]
  - gradual changes in intercepts, or time-lagged effects over time  

- Conceptually different: **Moderated VAR models** [@Holtmann2025] as a multilevel extension of single-subject fixed moderated time series analysis [@Adolf2017]
  
  - Moderation effects on the dynamic within-level where the moderator is known 
  - Inclusion of time-varying variables as moderators of dynamic regression effects (i.e., *AR* and *CL*)

**Possible research questions**

- Does engaging in emotion regulation strategies (e.g., rumination) at time $t-1$ impact the carry-over of negative affect at $t-1$ to the subsequent time point (i.e., moderated *AR*)? 


---

### A Simple Example

**Do positive affect dynamics change from the baseline to the intervention phase?** 

:::{.vspace}
:::

:::: {.columns}

::: {.column width="45%"}

- For simplicity, we assume a time-varying binary indicator variable for the study phases (i.e. `phase_bi`)

- We want `phase_bi` to be treated as an exogenous variable:
  (1) it only serves as predictor
  (2) without latent-mean centering 


:::

::: {.column width="5%"}
:::

::: {.column width="50%"}

#### Decomposition

- Latent Person-mean-centering only for positive affect ($PA$)

$$
PA_{it}^{w} = PA_{it} - \mu_{PA}
$$ 

#### Within-level
$$
PA_{it}^w = \phi_{21i} Phase_{it} + \phi_{22i} PA_{i(t-1)}^{w} + \phi_{2.21i} PA_{i(t-1)}^{w}*Phase_{it} + \zeta_{it} \\
$$

Implies:
$$
\begin{matrix}
\widehat{PA_{it}^w} &=& \phi_{22i} PA_{i(t-1)}^{w}  & \text{if}\ Phase_{it} = 0 \\
\widehat{PA_{it}^w} &=& \phi_{21i} + (\phi_{22i}+\phi_{2.21i})PA_{i(t-1)}^{w} & \text{if}\ Phase_{it} = 1\\
\end{matrix}
$$

$$
\text{with}\ \zeta_{it} \sim N(0, \sigma_{\zeta i}^2)
$$ 


:::

::::

--- 

### We build the model in a step-wise manner

::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  # two constructs: 1 = phase_bi, 2 = posaff
  q = 2
  )
```

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::


--- 

### We build the model in a step-wise manner

::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  # two constructs: 1 = phase_bi, 2 = posaff
  q = 2,                          
  # treat first construct as exogenous:
  is_exogenous = 1
  )
```

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::

--- 

### We build the model in a step-wise manner


::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 2,                          
  is_exogenous = 1,               
  
  # PA (t) regressed on phase_bi (t):
  incl_t0_effects = "phi(0)_21",  

  # additional constraints: 
  ## remove default first-order CL effect
  fixef_zero = "phi(1)_21",       
  ## for technical reasons, set
  inno_covs_zero = TRUE           
  )
```

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::

--- 

### We build the model in a step-wise manner

::: {.column width="35%"}

#### Model Specification
```{r}
#| echo: true
model <- mlts_model(
  q = 2,                          
  is_exogenous = 1,               
  incl_t0_effects = "phi(0)_21",  
  fixef_zero = "phi(1)_21",       
  inno_covs_zero = TRUE,            
  
  # include interaction effect:
  incl_interaction_effects = "phi(i)_2.2(1)1(0)" 
  )
```


#### Breakdown of `phi(i)_2.2(1)1(0)`

$$
\begin{array}{lll}
\text{phi(i)_} &\longrightarrow& \text{interaction effect} \\
\text{2.}      &\longrightarrow& \text{DV (PA)} \\
2(1)           &\longrightarrow& \text{IV I (PA) at lag 1} \\ 
1(0)           &\longrightarrow& \text{IV II (SP) at lag 0} \\
\end{array}
$$

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 3.5, width = 6, cex_b = 0.7, 
           scale_between = 0.35,
           scale_within = 0.29,
           cex_w = 0.8,
           asp_w_b = 0.4,
           scale_within_inno = 0.3,
           cex_title = 0.8,
           w_y_offset = c(0.25, -0.25), 
           y_ind_labs = c("SP","PA"),
           y_fac_labs = c("SP", "PA"),
           remove_lag_lab = T
           )
```

:::

---

### Multiple-Indicator Factor Models

- DSEM model are oftentimes run for aggregate scores 
  - Composite across multiple ESM items 
  - For example, *happy*, *relaxed*, and *cheerful* as indicators of positive affect
  - Measurement error in observed variables is not accounted for

- Consequences of disregarding measurement error
  - For time series variables, detrimental effects on the reliability of (individual) lagged effect estimates [e.g., @duReliabilitiesIntraindividualVariability2018]   
  - Underestimated AR effects in AR models [e.g., @schuurmanIncorporatingMeasurementError2015]
  - Potentially distorted estimates of CL relations in VAR models [e.g., @schuurmanMeasurementErrorPersonspecific2019]
  - Low reliability of person-specific estimates can impact their use as predictors 

- Solution:  
  - including a measurement model (both at the between- and/or the within-level) and
  - modeling the within-person dynamics for the underlying latent factors
  - requires that multiple parallel indicators of the same construct are available

---

### Multiple-Indicator Factor Models


:::: {.columns}

::: {.column width="35%"}

#### Decomposition
$$
\begin{matrix}
Y_{1.1it}^{w} &= Y_{1.1it} - \mu_{1.1i} \\
[. . .] \\
Y_{q.pit}^{w} &= Y_{q.pit} - \mu_{q.pi} \\
\end{matrix}
$$ 

#### Within-level measurement model
$$
\begin{matrix}
Y_{q.1it}^w &= \lambda_{q.p}^w \eta_{qit}^{w} + \varepsilon_{qit}^w \\
\end{matrix} \\
\text{with}\ \varepsilon_{qit}^w \sim N(0, \sigma_{\varepsilon_w,q.p}^2)
$$
For model identification: $\lambda_{q.1}^w = 1$.

#### Dynamic model (here AR(1))
$$
\begin{matrix}
\eta_{qit}^{w} &= \phi_{11i} \eta_{qi(t-1)}^{w} + \zeta_{qit} \\
\end{matrix} \\
\text{with}\ \zeta_{1it} \sim N(0, \sigma_{\zeta 1i}^2) \\
$$

:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

#### Model Specification

```{r, echo = T}
model = mlts_model(
  q = 1,              # number of time-varying constructs
  p = 3,              # number of manifest indicator
  btw_factor = FALSE, # no common between-level factor 
  )
```


```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 4, width = 6, cex_b = 0.8, 
           asp_decomp = 0.4,
           scale_within_inno = 0.22,
           cex_title = 0.8, scale_between = 0.38, asp_w_b = 0.4)
```
:::
::::


---

### Multiple-Indicator Factor Models

:::: {.columns}

::: {.column width="35%"}

#### Decomposition
$$
\begin{matrix}
Y_{1.1it}^{w} &= Y_{1.1it} - \mu_{1.1i} \\
[. . .] \\
Y_{q.pit}^{w} &= Y_{q.pit} - \mu_{q.pi} \\
\end{matrix}
$$ 


#### Between-level (indicator-specific person means)
$$
\begin{matrix}
\mu_{1.1i}  = \gamma_{\mu1.1}  + \upsilon_{\mu1.1i} \\
[...] \\
\mu_{q.pi}  = \gamma_{\mu q.p}  + \upsilon_{\mu q.pi} \\
\phi_{11i} = \gamma_{\phi11} + \upsilon_{\phi 11i} \\
ln(\sigma_{\zeta 2i}^2) = \gamma_{ln(\sigma^2_{\zeta2})} + \upsilon_{ln(\sigma^2_{\zeta 2i})}
\end{matrix} \\
\text{with}\ \mathbf{\upsilon} \sim MVN(\mathbf{0},\mathbf{\Sigma})
$$ 
:::

::: {.column width="5%"}
<!-- Add some space between columns -->
:::

::: {.column width="60%"}

#### Model Specification

```{r, echo = T}
model = mlts_model(
  q = 1,              # number of time-varying constructs
  p = 3,              # number of manifest indicator
  btw_factor = FALSE, # no common between-level factor 
  )
```


```r
mlts_paths(model)
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
#| fig-dpi: 1000
#| fig-align: left

mlts_paths(model, height = 4, width = 6, cex_b = 0.8, 
           asp_decomp = 0.4,
           scale_within_inno = 0.22,
           cex_title = 0.8, scale_between = 0.38, asp_w_b = 0.4)
```
:::
::::

---

### Extended modeling options

::: {.column width="50%"}

#### Common betwen-level factor 

```{r}
#| echo: true
model = mlts_model(
  q = 1, 
  p = 3, 
  btw_factor = TRUE)
```

```{r}
#| fig-width: 5
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.4, cex_b = 0.9)
```

:::

  
::: {.column width="50%"}

#### Complex VAR(1) structure 

```{r}
#| echo: true
model = mlts_model(
  q = 2, 
  p = c(3,4), 
  btw_model = list(1:3, NA),
  fix_dynamics = T,
  fix_inno_vars = T)
```

```{r}
#| fig-width: 7
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.6,
           cex_w = 0.7)
```

:::  
  
---  
  
### Extended modeling options 
  
::: {.column width="50%"}

#### Assuming cross-level measurement invariance 

```{r}
#| echo: true
model = mlts_model(
  q = 1, 
  p = 3, 
  btw_factor = TRUE, 
  equal_loads_levels = T
  )
```

```{r}
#| fig-width: 5
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.4, cex_b = 0.9)
```

:::
  
  
::: {.column width="50%"}

#### Additional constraints

```{r}
#| echo: true
model = mlts_model(q = 1, p = 3)

# Fix loading parameter to 1
model$Constraint[model$Param == "lambdaW_1.2"] <- "= 1"
# Fix error variance to be 0
model$Constraint[model$Param == "sigmaB_1.2"] <- "= 0"
# Remove constraints (agian)
model$Constraint[model$Param == "sigmaB_1.2"] <- "free"

```

```{r}
#| fig-width: 5
#| fig-height: 4
mlts_paths(model = model, width = 5, height = 4, asp_decomp = 0.45, cex_b = 0.9)
```

:::    

--- 


### Censoring Models

- Approach to modeling ESM data with floor or ceiling effects
- Floor effects challenge the normality assumptions underlying standard DSEM models 
- Specific assumption about the data generating process underlying floor effects 
  - Measurement-related censoring (i.e., observations are censored during measurement)
  - Implies: There is substantial variation in the underlying construct not captured by the item that exhibits floor effects
  
#### Evidence in favor of the censoring notion?

::: {.column width="60%"}
```{r}
#| fig-width: 12
#| fig-height: 5
#| fig-pos: left

# describe_ts(data = bolzen, ts = c("nervoes", "entspannt"), group = "id") %>%
#   arrange(desc(floor_pc))

bolzen %>%
  filter(id %in% c(11, 35)) %>%
  group_by(id) %>%
  mutate(
    entspannt_r = 8-entspannt
  ) %>%
  
  ggplot(aes(y = nervoes, x = obs_counter)) +
  geom_line() +
  geom_line(aes(y = entspannt_r), color = "darkgreen") +
  geom_point(alpha = 0.5) +
  geom_point(aes(y = entspannt_r), color = "darkgreen", alpha = 0.5) +
  geom_text(x = 87, y = 1.2, hjust = 0, label = "not nervous") +
  geom_text(x = 87, y = 1.9, hjust = 0, label = "relaxed",color = "darkgreen") +
  geom_text(x = 87, y = 6.8, hjust = 0, label = "nervous") +
  geom_text(x = 87, y = 6.1, hjust = 0, label = "not relaxed",color = "darkgreen") +
  scale_x_continuous(limits = c(0,95), expand = c(0,1)) +
  scale_y_continuous(limits = c(1,7), breaks = 1:7) +
  labs(x = "# Beep", y = NULL) +
  facet_grid(rows = vars(id)) +
  theme_classic() +
  theme(
    panel.spacing.y = unit(0.5, "cm")
  )


```
:::

::: {.column width="40%"}

- Inspecting unipolar items capturing the opposing ends of a continuum can be informative
- Here, as illustrated for two individuals, when subjects report to feel not nervous at all, they still show substantial variation in the relaxed item in these moments
- Hence, on the level of the broader construct, this suggests at qualitative differences in states of tense arousal beyond what is captured by the nervous item


:::


--- 

### Censoring Models

::: {.column width="80%"}

#### Impact of Ignoring Censoring-Related Floor Effects 

- Introduces spurious positive dependencies between person-specific estimates 
  - mean-variance-relation(s) most affected
- Censoring on the level of individual ESM items can affect estimates of analyses that use composite scores across indicators that exhibit floor effects [@Mader2023]
  - Composite scores often show more desirable distributional properties (e.g., less floor effects)
  - However, adverse effects on parameter estimates are not fully resolved by item aggregation
  - → Censored multiple-indicator factor models [@Koslowski2026]

:::


---

### Censoring Models

#### Implementing Censoring Models 

::: {.columns}
::: {.column width="47%"}

- Any model can be extended to accommodate censored variables   
- For example, a censored AR(1):

```r
mlts_model(q = 1, censor_left = 1)
```

- where `censor_left = 1` indicates all observations $\leq$ 1 will be assumed as censored
- These observations will be replaced with range-restricted latent parameters during model estimation 
  - Think of it as missing value imputation with the restriction that imputed values can not exceed the censoring threshold
  
:::

::: {.column width="6%"}

:::

::: {.column width="47%"}

:::{.vspace}
:::

:::{.vspace}
:::

#### Caveats 

- Censoring threshold can only be specified globally
  - ergo ALL variables will be affected by censoring
- In multivariate models, when this is not intended, we can use a workaround
  - Assume two variables measured on a scale of 1 to 7
  - To only treat one variable as censored, first shift the scale of that variable
  - Then use the new lower scale bound as the censoring threshold 

```r
Y1 = Y1 - 0.1   # shifts the lower scale end to 0.9

mlts_model(q = 2, censor_left = 0.9) 
```

:::
:::


